---
title: "Gulf of Naples - 40 Years of Zooplankton Biodiversity Assessment (ZOOGoN-40Y)"
author: "Iole di Capua (iole.dicapua@szn.it), Andrea Marcucci (andrea.marcucci@szn.it), Lorenzo Petrella (lorenzo.petrella@szn.it), Lorenzo Longobardi (lorenzo.longobardi@szn.it)"
date: today
format:
  html:
    theme: Darkly
    toc: true
    number-sections: true
    code-fold: true
    code-tools: true
    self-contained: true
  pdf:
    toc: true
    number-sections: true
---

# Aim
This report presents a spatio-temporal analysis of zooplankton abundance patterns at the MareChiara Long-Term Ecological Research (LTER) station in the Gulf of Naples, Mediterranean Sea, covering data recorded between 1984 and 2024. MareChiara is one of the longest-running marine monitoring stations in the Mediterranean, providing a rare 40-year window into plankton community dynamics.

The analysis employs two complementary visualization approaches to characterize temporal patterns and detect long-term trends:

1. Heatmap visualization: a multi-panel display showing weekly abundance patterns across years, integrated with seasonal profiles and annual anomalies to identify recurrent cycles and interannual variability

2. Trend analysis: a quarterly-stratified approach using anomaly-based regression models to isolate and quantify long-term trends while accounting for natural seasonal variability.

These methods enable the detection of phenological shifts, regime changes, and seasonal-specific trends in copepod abundance over the 40-year monitoring period.

# Methods
# Data Import and manipulation 

Data were obtained from the "Gulf of Naples - 40 Years of Zooplankton Biodiversity Assessment" project, funded by Flanders Marine Institute (VLIZ) and  Stazione Zoologica Anton Dohrn (SZN). Three temporal datasets were developd from four decades of sampling: McZoo_84-15.parquet, McZoo_16-20.parquet, and McZoo_21-24.parquet, covering the periods 1984-2015, 2016-2020, and 2021-2024, respectively.

Each dataset contains taxonomic identification, life stage classification, and abundance values (ind m⁻³) for zooplankton taxa. The three datasets were merged into a single time series, and copepod records were filtered for analysis. 

A methodological change occurred in 2021 with the adoption of KoBoToolbox as the data collection platform. This transition introduced the `isCopepod` variable (yes/no) in the 2021-2024 dataset to facilitate taxonomic filtering. To ensure consistency across the entire time series, the variables defining the taxonomic classification have also been included in previous datasets (1984-2020) using the `worrms` package. Data manipulation and summarization were performed using the `dplyr` package, while visualizations were created with `ggplot2`, `cowplot`, and `viridis` packages.

The code below demonstrates the data loading process and displays the first 10 rows of the final dataset after merge with WoRMS taxonomic data:

```{r}
#| echo: true
#| code-fold: true
# require needed packages
pacman::p_load(
  magrittr,
  scales,
  ggsci,
  lubridate,
  tidyverse,
  viridis,
  cowplot,
  worrms,
  stringr,
  vegan,
  arrow
)

# load package and configurations
if (requireNamespace("ZooGoN", quietly = TRUE)) {
  library(ZooGoN)
} else {
  devtools::load_all()
}
conf <- read_config()

# load legacy data from SharePoint
legacy_84_15 <- ZooGoN::download_sharepoint_file(
  prefix  = "McZoo_84-15.parquet",
  options = conf$storage$sharepoint$credentials,
  bucket  = conf$storage$sharepoint$buckets$hot_bucket,
  filename = TRUE
)

legacy_16_20 <- ZooGoN::download_sharepoint_file(
  prefix  = "McZoo_16-20.parquet",
  options = conf$storage$sharepoint$credentials,
  bucket  = conf$storage$sharepoint$buckets$hot_bucket,
  filename = TRUE
)

# load processed survey data from pipeline (2021-24 + futuri)
legacy_21_24 <- ZooGoN::download_sharepoint_file(
  prefix  = conf$ingestion$surveys$preprocessed$file_prefix,
  options = conf$storage$sharepoint$credentials,
  bucket  = conf$storage$sharepoint$buckets$automation_bucket,
  format  = "parquet"
)

# merge into a single databse
legacy_84_24 <-
  dplyr::bind_rows(legacy_84_15, legacy_16_20, legacy_21_24)

# add Class and Order from Worms
# extrac AphiaID to lsid
aphiaid <- 
  legacy_84_24 |>
  dplyr::mutate(
    aphiaID = as.integer(stringr::str_extract(lsid, "[0-9]+"))
  )

# Extract unique AphiaIDs to avoid duplicate WoRMS API calls
ids <- unique(aphiaid$aphiaID)

# get WORms taxonomy
get_class_order <- function(id) {
  tryCatch(
    {
      cls <- wm_classification(id)
      # extract scientificname where rank matches, return NA if not found
      class_val <- cls$scientificname[cls$rank == "Class"]
      order_val <- cls$scientificname[cls$rank == "Order"]
      data.frame(
        aphiaID  = id,
        class = if (length(class_val) > 0) class_val[1] else NA_character_,
        order = if (length(order_val) > 0) order_val[1] else NA_character_,
        stringsAsFactors = FALSE
      )
    },
    error = function(e) {
      data.frame(
        aphiaID  = id,
        class = NA_character_,
        order = NA_character_,
        stringsAsFactors = FALSE
      )
    }
  )
}

ranks_df <- bind_rows(lapply(ids, get_class_order))

# merge
zoo_84_24 <-
  aphiaid |>
  dplyr::left_join(ranks_df, by = "aphiaID") |>
  dplyr::select(-aphiaID) |>
  dplyr::relocate(class, order, .before = scientificName) |>
  dplyr::rename(rankCalss = class, rankOrder = order)

head(zoo_84_24, 10)
```


Prior to visualization, the dataset was restructured to optimize plotting performance. The `cop_heatmap` dataset aggregates abundance values by week and year, and includes a pre-computed log-transformed variable for species-specific analyses. 

The code shows the first 10 rows of the final dataset: 
```{r}
#| echo: true
#| code-fold: true
zoo_heatmap <-
  zoo_84_24 |>
  dplyr::mutate(
    week = lubridate::week(eventDate), # Extract week (1-52)
    year = lubridate::year(eventDate) # Extract year
  ) |>
  dplyr::group_by(year, week, rankCalss, rankOrder, scientificName) |> # Group by year, week, species and life stage
  dplyr::summarise(
    abundance = sum(individualCount, na.rm = TRUE),
    .groups = "drop"
  ) |> # Sum abundances
  dplyr::mutate(log_abundance = log(abundance + 1)) # Log-transformation (+1 to avoid log(0))

head(zoo_heatmap, 10)
```

In `plot_cop_heatmap_complex()` abundance values are log-transformed by default (log = TRUE) in order to reduce skewness and improve the visualization of low and high abundance values.The log-transformation can be disabled by setting log = FALSE. The function function generates a three-panel display showing weekly abundance patterns across years, seasonal profiles, and annual anomalies.

See the code comments below:

```{r}
#| echo: true
#| code-fold: true
# heatmap function
plot_heatmap_complex <- function(
  species_name = NULL,
  rank_class = NULL,
  rank_order = NULL,
  log = TRUE
) {
  # apply taxonomic filters first (if provided), then aggregate or filter by species
  data <-
    zoo_heatmap |>
    dplyr::filter(
      if (!is.null(rank_class)) rankCalss == rank_class else TRUE, # filter by class if provided
      if (!is.null(rank_order)) rankOrder == rank_order else TRUE, # filter by order if provided
      if (!is.null(species_name) && species_name != "All") {
        scientificName == species_name
      } else {
        TRUE
      } # filter by species if provided
    )

  # aggregate all species if species_name is NULL or "All"
  if (is.null(species_name) || species_name == "All") {
    data <-
      data |>
      dplyr::group_by(year, week) |>
      dplyr::summarise(
        abundance = sum(abundance, na.rm = TRUE),
        .groups = "drop"
      ) |>
      dplyr::mutate(
        abundance_plot = if (log) log(abundance + 1) else abundance
      )
  } else {
    # filter by specific species
    data <-
      data |>
      dplyr::mutate(
        abundance_plot = if (log) log_abundance else abundance
      )
  }

  # build dynamic title based on what is being plotted
  plot_title <- if (!is.null(species_name) && species_name != "All") {
    species_name
  } else if (!is.null(rank_order)) {
    rank_order
  } else if (!is.null(rank_class)) {
    rank_class
  } else {
    "All Zooplankton"
  }

  # legend text and theme
  legend_text <- expression(bold(ind ~ m^-3))

  legend_theme <- ggplot2::theme(
    legend.position = "bottom",
    legend.key.width = unit(0.5, "cm"),
    legend.direction = "horizontal",
    legend.key.size = unit(0.2, "cm"),
    legend.title = ggplot2::element_text(size = 10, face = "bold"),
    legend.text = ggplot2::element_text(size = 10, face = "bold")
  )

  # dynamic breaks: visually centered on pseudo-log scale regardless of data range
  min_val <- min(data$abundance_plot[data$abundance_plot > 0], na.rm = TRUE)
  max_val <- max(data$abundance_plot, na.rm = TRUE)

  # compute midpoint on the transformed scale, then back-transform
  sigma <- 5
  mid_transformed <- (asinh(min_val / sigma) + asinh(max_val / sigma)) / 2
  mid_val <- round(sinh(mid_transformed) * sigma) # back-transform to original scale

  breaks_vals <- c(0, mid_val, round(max_val))

  # main heatmap: weeks (x) vs years (y), colored by abundance
  heatmap_plot <-
    data |>
    ggplot2::ggplot(ggplot2::aes(x = week, y = year, fill = abundance_plot)) +
    ggplot2::geom_tile(color = "grey90") +
    ggplot2::theme_bw() +
    ggplot2::scale_fill_gradientn(
      colours = c("blue", "cyan", "green", "yellow", "orange", "red"),
      trans = scales::pseudo_log_trans(sigma = 5),
      breaks = breaks_vals, # dynamic breaks adapted to the plotted data
      labels = scales::label_number(),
      na.value = "grey90"
    ) +
    ggplot2::coord_cartesian(expand = FALSE) +
    ggplot2::scale_x_continuous(breaks = seq(1, 52, by = 5)) +
    ggplot2::scale_y_continuous(breaks = seq(1984, 2024, by = 4)) +
    ggplot2::theme(
      panel.background = ggplot2::element_blank(),
      panel.grid = ggplot2::element_blank(),
      legend.box.just = "left",
      legend.box.margin = margin(t = 0, r = 300, b = 0, l = 0),
      axis.text.x = ggplot2::element_text(size = 12, face = "bold"),
      axis.text.y = ggplot2::element_text(size = 12, face = "bold"),
      axis.title.x = ggplot2::element_text(size = 16, face = "bold"),
      axis.title.y = ggplot2::element_text(size = 16, face = "bold"),
      plot.title = ggplot2::element_text(size = 16, face = "bold")
    ) +
    legend_theme +
    ggplot2::labs(
      x = "Weeks",
      y = "Years",
      fill = legend_text
    ) +
    guides(alpha = "none")

  # top marginal plot: mean abundance by week across all years
  line <-
    data |>
    dplyr::group_by(week) |>
    dplyr::summarise(abundance_plot = mean(abundance_plot, na.rm = TRUE))

  # identify min and max values for annotation
  minmax <-
    line |>
    dplyr::mutate(
      label_row = dplyr::case_when(
        abundance_plot == min(abundance_plot, na.rm = TRUE) ~ "min",
        abundance_plot == max(abundance_plot, na.rm = TRUE) ~ "max",
        TRUE ~ NA_character_
      )
    ) |>
    dplyr::filter(!is.na(label_row)) |>
    dplyr::arrange(abundance_plot)

  line_plot <- ggplot2::ggplot() +
    ggplot2::theme_void() +
    ggplot2::stat_smooth(
      data = line,
      ggplot2::aes(week, abundance_plot),
      method = "loess",
      color = "#d64d4d",
      se = FALSE,
      span = 0.25,
      linewidth = 0.5
    ) +
    ggplot2::geom_area(
      data = line,
      ggplot2::aes(week, abundance_plot),
      fill = "grey95",
      color = "grey70"
    ) +
    ggplot2::geom_point(
      data = line,
      ggplot2::aes(week, abundance_plot, color = abundance_plot)
    ) +
    ggplot2::geom_text(
      data = minmax,
      ggplot2::aes(
        x = week,
        y = abundance_plot,
        label = round(abundance_plot, 1),
        vjust = ifelse(label_row == "max", 1.5, -0.5)
      ),
      color = "firebrick",
      size = 6
    ) +
    ggplot2::geom_point(
      data = minmax,
      ggplot2::aes(x = week, y = abundance_plot),
      color = "firebrick"
    ) +
    ggplot2::scale_color_gradientn(
      colours = c("blue", "cyan", "green", "yellow", "orange", "red")
    ) +
    ggplot2::scale_x_continuous(limits = c(1, 52), expand = c(0, 0)) +
    ggplot2::coord_cartesian(xlim = c(1, 52), expand = FALSE, clip = "off") +
    ggplot2::theme(
      legend.position = "none",
      plot.margin = margin(t = 15, b = 5, l = 70, r = 69)
    )

  # right side panel: annual anomalies as % deviation from mean
  side_tile <-
    data |>
    dplyr::group_by(year) |>
    dplyr::summarise(abundance_plot = mean(abundance_plot, na.rm = TRUE)) |>
    dplyr::mutate(
      mean_abundance = mean(abundance_plot, na.rm = TRUE),
      mean_anomaly = (abundance_plot - mean_abundance) / mean_abundance * 100,
      mean_anomaly_scale = pmax(pmin(mean_anomaly, 100), -100)
    )

  side_plot <-
    side_tile |>
    ggplot2::ggplot() +
    ggplot2::geom_tile(
      ggplot2::aes(x = "group", y = year, fill = mean_anomaly_scale),
      alpha = 0.6,
      color = "grey90"
    ) +
    ggplot2::geom_text(
      ggplot2::aes(
        x = "group",
        y = year,
        label = ifelse(
          !is.na(mean_anomaly),
          paste0(round(mean_anomaly, 0), "%"),
          "-"
        )
      ),
      size = 2,
      color = "grey30"
    ) +
    ggplot2::theme_void() +
    ggplot2::scale_fill_gradient2(
      low = "#56a8a9",
      mid = "white",
      high = "#a95756",
      midpoint = 0,
      na.value = "white"
    ) +
    ggplot2::scale_y_continuous(breaks = seq(1984, 2024, by = 4)) +
    ggplot2::coord_cartesian(expand = FALSE) +
    ggplot2::labs(fill = "Mean difference (%)") +
    ggplot2::theme(
      plot.margin = margin(t = 0, r = 12, b = 0, l = -3.5),
      panel.background = ggplot2::element_blank(),
      panel.grid = ggplot2::element_blank(),
      legend.box.just = "right",
      legend.box.margin = margin(t = 0, r = 0, b = 0, l = -250)
    ) +
    legend_theme +
    guides(alpha = "none")

  # combine panels
  p1 <- cowplot::plot_grid(
    heatmap_plot,
    side_plot,
    ncol = 2,
    rel_widths = c(12, 1),
    align = "h"
  )

  combined <- cowplot::plot_grid(line_plot, p1, nrow = 2, rel_heights = c(1, 5))

  # add title
  title <- cowplot::ggdraw() +
    cowplot::draw_label(
      bquote(italic(.(plot_title))), # dynamic title based on what is being plotted
      fontface = 'bold',
      size = 20,
      x = 0,
      hjust = 0
    ) +
    ggplot2::theme(
      plot.margin = margin(0, 0, 0, 7)
    )

  # return final plot
  cowplot::plot_grid(
    title,
    combined,
    ncol = 1,
    rel_heights = c(0.05, 1)
  )
}
```

The `plot_temporal_trend()` function performs deseasonalized anomaly analysis with quarterly stratification to detect long-term trends.

See the code and comments below:
```{r}
#| echo: true
#| code-fold: true
# trend function
plot_temporal_trend <- function(species_name = NULL, rank_class = NULL, rank_order = NULL) {

  # apply taxonomic filters first (if provided)
  zoo_filtered <-
    zoo_84_24 |>
    dplyr::filter(
      if (!is.null(rank_class)) rankCalss == rank_class else TRUE,   # filter by class if provided
      if (!is.null(rank_order)) rankOrder == rank_order else TRUE,   # filter by order if provided
      if (!is.null(species_name) && species_name != "All") scientificName == species_name else TRUE  # filter by species if provided
    )

  # aggregate all species if species_name is NULL or "All"
  if (is.null(species_name) || species_name == "All") {
    filtered_data <-
      zoo_filtered |>
      dplyr::group_by(eventDate) |>
      dplyr::summarise(
        abundance = sum(individualCount, na.rm = TRUE),
        .groups = "drop"
      )
  } else {
    filtered_data <-
      zoo_filtered |>
      dplyr::filter(scientificName == species_name) |>
      dplyr::group_by(eventDate) |>
      dplyr::summarise(
        abundance = sum(individualCount, na.rm = TRUE),
        .groups = "drop"
      )
  }

  # build dynamic title based on what is being plotted
  plot_title <- if (!is.null(species_name) && species_name != "All") {
    species_name
  } else if (!is.null(rank_order)) {
    rank_order
  } else if (!is.null(rank_class)) {
    rank_class
  } else {
    "All Zooplankton"
  }

  # calculate deseasonalized anomalies with quarterly stratification
  data_df <-
    filtered_data |>
    dplyr::mutate(
      month = lubridate::month(eventDate),
      quarter = lubridate::quarter(eventDate), # Q1-Q4 for seasonal analysis
      year = lubridate::year(eventDate),
      month_date = lubridate::floor_date(eventDate, "month"),
      series = ifelse(year < 1993, "before", "after") # temporal split
    ) |>
    dplyr::group_by(series, month) |>
    dplyr::mutate(
      mean_abundance = mean(abundance),
      anomaly = abundance - mean_abundance # remove seasonal cycle
    ) |>
    dplyr::group_by(series, month_date) |>
    dplyr::summarise(
      quarter = dplyr::first(quarter),
      month = dplyr::first(month),
      anomaly = mean(anomaly),
      .groups = "drop"
    ) |>
    dplyr::mutate(
      quarter = as.character(quarter)
    ) |>
    dplyr::ungroup()

  # split into before/after periods
  data_df <- split(data_df, data_df$series)

  # linear regression by quarter (post-1993 only)
  trend_res <-
    data_df$after |>
    dplyr::group_by(quarter) |>
    dplyr::do(broom::tidy(lm(anomaly ~ month_date, data = .))) |>
    dplyr::filter(term == "month_date") |>
    dplyr::select(quarter, estimate, p.value)

  # format statistical results as text annotations
  annotations_text <-
    trend_res |>
    dplyr::arrange(quarter) |>
    dplyr::mutate(
      label = sprintf("Q%s: %.4f, p=%.4f", quarter, estimate, p.value)
    ) |>
    dplyr::pull(label) |>
    paste(collapse = "\n")

  # time series plot with quarterly color coding
  ggplot2::ggplot() +
    ggplot2::theme_minimal() +
    # points for "before" period (pre-1993)
    ggplot2::geom_point(
      data = data_df$before,
      ggplot2::aes(month_date, anomaly, color = quarter, alpha = abs(anomaly))
    ) +
    # points for "after" period (post-1993)
    ggplot2::geom_point(
      data = data_df$after,
      ggplot2::aes(month_date, anomaly, color = quarter, alpha = abs(anomaly))
    ) +
    # connecting lines
    ggplot2::geom_line(
      data = data_df$before,
      ggplot2::aes(month_date, anomaly),
      alpha = 0.3,
      linewidth = 0.1
    ) +
    ggplot2::geom_line(
      data = data_df$after,
      ggplot2::aes(month_date, anomaly),
      alpha = 0.3,
      linewidth = 0.1
    ) +
    # zero reference line (climatological mean)
    ggplot2::geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
    # linear trend lines by quarter (post-1993 only)
    ggplot2::stat_smooth(
      data = data_df$after,
      ggplot2::aes(month_date, anomaly, group = quarter, color = quarter),
      method = "lm",
      se = FALSE,
      linewidth = 0.8
    ) +
    # JAMA color palette for quarterly distinction
    ggsci::scale_color_jama() +
    ggplot2::scale_x_date(
      date_breaks = "4 years",
      labels = scales::label_date(format = "%b\n%Y", locale = "en_US.UTF-8")
    ) +
    ggplot2::coord_cartesian(expand = FALSE) +
    # statistical results annotation
    ggplot2::annotate(
      "text",
      x = as.Date("1985-12-01"),
      y = max(data_df$after$anomaly, na.rm = TRUE),
      label = annotations_text,
      hjust = 0,
      vjust = 1,
      color = "grey40"
    ) +
    ggplot2::theme(legend.position = "bottom") +
    ggplot2::guides(alpha = "none") +
    ggplot2::labs(
      title = bquote(italic(.(plot_title))),  # dynamic title based on what is being plotted
      x = "Date",
      y = "Monthly mean deviation\n(Abundance)",
      color = "Annual quarter"
    )
}
```

# Results
# Heatmap
The heatmap plots illustrates the abundance across different weeks and years. The x-axis represents the weeks of the year, ranging from 1 to 52, while the y-axis spans several years. Each cell in the heatmap indicates the abundance in individuals per cubic metre, with a color gradient from blue (low abundance) to red (high abundance), as indicated by the legend at the bottom. At the top of the heatmap, a line plot displays the average weekly abundance over the entire period, highlighting the seasonality of abundance. A red point marks the mean maximum abundance, while another red point indicates the first non-zero mean value, providing insight into the average abundance range of the species. On the right side of the heatmap, a side plot displays the mean annual deviation in percentage from the overall annual mean.

# Temporal trend
The temporal trend plot illustrates the mean deviation in abundance over time, featuring a linear regression line and the p-value of the regression for each quarter of the year. The x-axis represents the date, and the y-axis shows the monthly mean deviation in abundance. This deviation is calculated by subtracting the overall mean for each specific month from the corresponding monthly values in the series. For example, each January value is subtracted from the overall January mean, each February value from the overall February mean, and so on.

# Data Visualization Examples 
The following code illustrates how to call the two main visualization functions at different taxonomic levels: 
```{r}
#| echo: true
#| code-fold: true
plot_heatmap_complex(species_name = "All", log = F) # to show all zooplankton 
plot_heatmap_complex(rank_class = "Copepoda", log = F) # to show a single class
plot_heatmap_complex(rank_order = "Onychopoda", log = F) # to show a single order 
plot_heatmap_complex(species_name = "Acartia clausi", log = F) # to show single sepcies 
```


```{r}
#| echo: true
#| code-fold: true 
plot_temporal_trend(species_name = "All") # to show all zooplankton
plot_temporal_trend(rank_class = "Thecostraca") # to show single class
plot_temporal_trend(rank_order = "Euphausiacea") # to show single order
plot_temporal_trend(species_name = "Acartia clausi") # to show single species
```

# Data Pipeline & Report Automation
# Data Sources

The three temporal datasets used in this report covering 1984–2015, 2016–2020, and 2021–present; are stored and accessed via SharePoint. The legacy datasets (1984–2020) were previously maintained on local computers and have been migrated to a centralized cloud storage to ensure reproducibility and accessibility. From 2021 onwards, data collection relies on [**KoBoToolbox**](https://www.kobotoolbox.org) digital survey forms, which replaced the previous local workflow. 

# Automated Processing Pipeline
Data are processed through the [**ZooGoN**](https://github.com/ioledc/ZOOGoN-40Y) R package, which implements a fully automated pipeline structured in two 
parallel way, both triggered daily via [**GitHub Actions**](https://github.com/ioledc/ZOOGoN-40Y/actions/runs/22932728351). 
The complete pipeline workflow below: 
```{mermaid}
%%| fig-width: 10
flowchart LR
    A[Build R Container] --> B[Ingest MC Kobo Surveys]
    B --> C[Process MC Kobo Surveys]
    C --> D[Merge Tidy Analysis-Ready Dataset]
    C --> E[Convert to Darwin Core Format]
    D --> F[Render MC Survey Report]
    E --> G[Build Darwin Core Archive]

    style A fill:#000000,color:#ffffff,stroke:#ffffff
    style B fill:#000000,color:#ffffff,stroke:#ffffff
    style C fill:#000000,color:#ffffff,stroke:#ffffff
    style D fill:#000000,color:#ffffff,stroke:#ffffff
    style E fill:#000000,color:#ffffff,stroke:#ffffff
    style F fill:#000000,color:#ffffff,stroke:#ffffff
    style G fill:#000000,color:#ffffff,stroke:#ffffff
```

# Analysis & Report
The function `ingest_surveys()` connects to the KoBoToolbox API, downloads new survey submissions with pagination support, flattens nested JSON data into tabular format, and archives the raw data on SharePoint as CSV and Parquet files. 
```{r}
#| echo: true
#| code-fold: true
ingest_surveys <- function() {
  conf <- read_config()

  logger::log_info("Downloading MC Survey Kobo data...", namespace = "ZooGoN")
  data_raw <-
    get_kobo_data(
      url = "eu.kobotoolbox.org",
      assetid = conf$ingestion$kobo$asset_id,
      uname = conf$ingestion$kobo$username,
      pwd = conf$ingestion$kobo$password,
      encoding = "UTF-8",
      format = "json"
    )

  # Check that submissions are unique in case there is overlap in the pagination
  if (
    dplyr::n_distinct(purrr::map_dbl(data_raw, ~ .$`_id`)) != length(data_raw)
  ) {
    stop("Number of submission ids not the same as number of records")
  }

  logger::log_info("Converting MC Survey Kobo data to tabular format...", namespace = "ZooGoN")
  raw_survey <-
    purrr::map(data_raw, flatten_row) %>%
    dplyr::bind_rows() %>%
    dplyr::rename(submission_id = "_id")

  logger::log_debug("Flattened survey: {nrow(raw_survey)} rows, {ncol(raw_survey)} columns", namespace = "ZooGoN")

  logger::log_info("Uploading raw survey data (CSV + Parquet)...", namespace = "ZooGoN")
  c("csv", "parquet") |>
    purrr::walk(
      ~ upload_sharepoint_df(
        data = raw_survey,
        prefix = conf$ingestion$surveys$raw$file_prefix,
        options = conf$storage$sharepoint$credentials,
        bucket = conf$storage$sharepoint$buckets$automation_bucket,
        format = .
      )
    )

  logger::log_success("ingest_surveys complete", namespace = "ZooGoN")
}
```

The Function `preprocess_surveys()` cleans and transforms raw survey data, computes abundance values (ind m⁻³) by sex and life stage, and harmonizes taxonomic identifiers against [**WoRMS**](https://www.marinespecies.org/) via AphiaID.
```{r}
#| echo: true
#| code-fold: true
preprocess_surveys <- function(raw_data = NULL) {
  conf <- read_config()

  logger::log_info("Downloading raw survey data...", namespace = "ZooGoN")
  raw_surveys <-
    download_sharepoint_file(
      prefix = conf$ingestion$surveys$raw$file_prefix,
      options = conf$storage$sharepoint$credentials,
      bucket = conf$storage$sharepoint$buckets$automation_bucket,
      format = "csv"
    ) |>
    # remove kobo metadata columns
    dplyr::select(
      -dplyr::starts_with("_"),
      -dplyr::starts_with("meta/"),
      -dplyr::all_of(c("formhub/uuid", "start", "end", "today"))
    )

  logger::log_debug("Raw surveys: {nrow(raw_surveys)} rows", namespace = "ZooGoN")

  logger::log_info("Extracting cruise info...", namespace = "ZooGoN")
  cruise_info <-
    raw_surveys |>
    dplyr::select("submission_id", !dplyr::starts_with("group_taxa")) |>
    dplyr::rename_with(~ stringr::str_remove(., "group_cruise/")) |>
    dplyr::rename_with(~ stringr::str_remove(., "group_environment/")) |>
    dplyr::rename_with(~ stringr::str_remove(., "group_cruise/")) |>
    dplyr::rename_with(~ stringr::str_remove(., "group_abundance/")) |>
    dplyr::rename_with(~ stringr::str_remove(., "group_sample/")) |>
    dplyr::rename(filtered_volume_m3 = "Filtered_Volume_m") |>
    dplyr::mutate(sampling_id = as.integer(.data$sampling_id)) |>
    #janitor::clean_names() |>
    dplyr::relocate(
      "filtered_volume_m3",
      .before = "water_column_sampled"
    ) |>
    # remove legacy columns
    dplyr::select(-"sampling_name")

  logger::log_info("Extracting taxa info...", namespace = "ZooGoN")
  taxa_info <-
    raw_surveys |>
    dplyr::select("submission_id", dplyr::starts_with("group_taxa")) |>
    reshape_kobo_repeat(group_name = "group_taxa") |>
    dplyr::mutate(
      taxon = dplyr::coalesce(
        .data$taxon_cope_sel,
        .data$taxon_noncope_sel
        # TO DO: fix missing other_taxon
        #.data$other_taxon
      )
    ) |>
    dplyr::select(
      -dplyr::all_of(c(
        "taxon_cope_sel",
        "taxon_noncope_sel"
        # TO DO: fix missing other_taxon
        #"other_taxon"
      ))
    ) |>
    dplyr::relocate("taxon", .after = "n_sample") |>
    # remove old duplicated labels
    dplyr::mutate(taxon = stringr::str_replace(.data$taxon, "_1", ""))

  preprocessed_survey <-
    dplyr::full_join(cruise_info, taxa_info, by = "submission_id") |>
    janitor::clean_names()

  # process abundances (ind_m3)
  preprocessed_complete <-
    preprocessed_survey |>
    dplyr::mutate(
      dplyr::across(
        .cols = c(
          "n_male",
          "n_female",
          "n_copepodite",
          "n_undetermined",
          "n_larvae",
          "n_eggs"
        ),
        ~ as.numeric(.x) *
          (.data$total_volume / .data$subsample_volume) /
          .data$filtered_volume_m3
      )
    )

  clean_21_24 <-
    preprocessed_complete |>
    dplyr::select(
      "site_name",
      "sampling_id",
      "sampling_date",
      "taxon",
      dplyr::starts_with("n_")
    ) |>
    dplyr::mutate(eventID = paste0(.data$site_name, .data$sampling_id)) |>
    dplyr::relocate("eventID", .before = "site_name") |>
    dplyr::select(-c("site_name", "sampling_id")) |>
    dplyr::arrange(.data$sampling_date) |>
    dplyr::rename(
      eventDate = "sampling_date",
      aphiaID = "taxon"
    ) |>
    janitor::clean_names() |>
    dplyr::mutate(dplyr::across(dplyr::starts_with("n_"), as.numeric)) |>
    # TO DO: Ideally there should be no NAs unless the net was empty(!), to clarify. Meanwhile we drop all NAs
    dplyr::filter(!is.na(.data$aphia_id))

  logger::log_debug("Cleaned data: {nrow(clean_21_24)} rows", namespace = "ZooGoN")

  # Get unique AphiaID
  unique_aphia_ids <- as.numeric(unique(clean_21_24$aphia_id))
  logger::log_info("Querying WoRMS for {length(unique_aphia_ids)} unique AphiaIDs...", namespace = "ZooGoN")

  worms_records <-
    unique_aphia_ids |>
    purrr::map_dfr(worrms::wm_record) |>
    dplyr::select(aphia_id = "AphiaID", "scientificname", "lsid") |>
    dplyr::mutate(aphia_id = as.character(.data$aphia_id))

  # Data harmonization as other matrices
  tidy_data <-
    clean_21_24 |>
    dplyr::left_join(worms_records, by = "aphia_id") |>
    dplyr::relocate("scientificname", "lsid", .after = "aphia_id") |>
    # Pivot abundance columns to long format
    tidyr::pivot_longer(
      cols = c(
        "n_male",
        "n_female",
        "n_copepodite",
        "n_undetermined",
        "n_larvae",
        "n_eggs",
        "n_nauplius"
      ),
      names_to = "life_stage_temp",
      values_to = "individual_count"
    ) |>
    # Convert abundance to numeric and map life stages
    dplyr::mutate(
      life_stage = dplyr::case_when(
        life_stage_temp == "n_male" ~ "m",
        life_stage_temp == "n_female" ~ "f",
        life_stage_temp == "n_copepodite" ~ "j",
        life_stage_temp == "n_undetermined" ~ "fmj",
        life_stage_temp == "n_larvae" ~ "lar",
        life_stage_temp == "n_eggs" ~ "egg",
        # TO DO: do we have nauplii only in 21-24 data?
        life_stage_temp == "n_nauplius" ~ "nau",
        TRUE ~ NA_character_
      )
    ) |>
    # Clean and organize columns
    dplyr::select(
      -c("life_stage_temp", "aphia_id", "n_individuals", "n_sample")
    ) |>
    dplyr::filter(!is.na(.data$individual_count)) |>
    dplyr::rename(
      scientificName = "scientificname",
      eventID = "event_id",
      eventDate = "event_date",
      lifeStage = "life_stage",
      individualCount = "individual_count"
    ) |>
    dplyr::distinct() |>
    # remove NA counts & IDs
    dplyr::filter(
      !is.na(.data$individualCount),
      !is.na(.data$eventDate),
      !is.na(.data$eventID)
    ) |>
    # remove duplicates
    dplyr::group_by(
      .data$eventID,
      .data$eventDate,
      .data$scientificName,
      .data$lsid,
      .data$lifeStage
    ) |>
    dplyr::summarise(
      individualCount = sum(.data$individualCount, na.rm = TRUE),
      .groups = "drop"
    ) |>
    dplyr::relocate("individualCount", .before = "lifeStage")

  logger::log_info("Preprocessed data: {nrow(tidy_data)} rows, {length(unique(tidy_data$scientificName))} unique taxa", namespace = "ZooGoN")
  logger::log_info("Uploading preprocessed data (CSV + Parquet)...", namespace = "ZooGoN")
  c("csv", "parquet") |>
    purrr::walk(
      ~ upload_sharepoint_df(
        data = tidy_data,
        prefix = conf$ingestion$surveys$preprocessed$file_prefix,
        options = conf$storage$sharepoint$credentials,
        bucket = conf$storage$sharepoint$buckets$automation_bucket,
        format = .
      )
    )
  logger::log_success("preprocess_surveys complete", namespace = "ZooGoN")
  # upload_sharepoint_df(
  #   data = preprocessed_complete,
  #   prefix = conf$ingestion$survey$preprocessed$file_prefix,
  #   options = conf$storage$sharepoint$credentials,
  #   bucket = conf$storage$sharepoint$buckets$automation_bucket,
  #   format = "csv"
  # )
}
```

The function `raw_to_tidy()` merges the legacy datasets (1984–2020) with the ongoing survey landings (2021–present) into a single long-format analysis-ready dataset, uploaded to SharePoint as CSV and Parquet.
```{r}
#| echo: true
#| code-fold: true
raw_to_tidy <- function() {
  conf <- read_config()

  logger::log_info("Downloading legacy datasets from hot storage...", namespace = "ZooGoN")
  legacy_84_20 <-
    c("McZoo_84-15.parquet", "McZoo_16-20.parquet") |>
    purrr::map(
      download_sharepoint_file,
      options = conf$storage$sharepoint$credentials,
      bucket = conf$storage$sharepoint$buckets$hot_bucket,
      filename = TRUE
    ) |>
    dplyr::bind_rows()
  logger::log_debug("Legacy data loaded: {nrow(legacy_84_20)} rows", namespace = "ZooGoN")

  logger::log_info("Downloading preprocessed ongoing surveys...", namespace = "ZooGoN")
  ongoing_landings <-
    download_sharepoint_file(
      prefix = conf$ingestion$surveys$preprocessed$file_prefix,
      options = conf$storage$sharepoint$credentials,
      bucket = conf$storage$sharepoint$buckets$automation_bucket,
      format = "parquet"
    )
  logger::log_debug("Ongoing surveys loaded: {nrow(ongoing_landings)} rows", namespace = "ZooGoN")

  tidy_data <-
    legacy_84_20 |>
    dplyr::bind_rows(ongoing_landings)
  logger::log_info("Merged tidy dataset: {nrow(tidy_data)} rows", namespace = "ZooGoN")

  logger::log_info("Uploading tidy data (CSV + Parquet)...", namespace = "ZooGoN")
  c("csv", "parquet") |>
    purrr::walk(
      ~ upload_sharepoint_df(
        data = tidy_data,
        prefix = conf$ingestion$tidy_data$file_prefix,
        options = conf$storage$sharepoint$credentials,
        bucket = conf$storage$sharepoint$buckets$automation_bucket,
        format = .
      )
    )
  logger::log_success("raw_to_tidy complete", namespace = "ZooGoN")
}
```

The function `render_report()` downloads the latest processed dataset from SharePoint and renders this report in HTML and PDF formats.
```{r}
#| echo: true
#| code-fold: true
render_report <- function(output_dir = "/home") {
  conf <- read_config()

  report_path <- system.file("report/REPORT.qmd", package = "ZooGoN")

  if (report_path == "") {
    stop("Report.qmd not found in inst/report/. Make sure the package is installed.")
  }

  logger::log_info("Rendering Quarto report...", namespace = "ZooGoN")
  quarto::quarto_render(
    input         = report_path,
    output_format = c("html", "pdf"),
    execute_params = list(
      sharepoint_site_url = conf$storage$sharepoint$credentials$site_url,
      data_prefix         = conf$ingestion$surveys$preprocessed$file_prefix,
      automation_bucket   = conf$storage$sharepoint$buckets$automation_bucket
    )
  )

  # Quarto saves files next to the .qmd — copy both to output_dir
  for (ext in c("html", "pdf")) {
    out_file <- paste0("REPORT.", ext)
    rendered_path <- file.path(dirname(report_path), out_file)
    dest_path <- file.path(output_dir, paste0("ZooGoN-report-", Sys.Date(), ".", ext))
    if (file.exists(rendered_path)) {
      file.copy(rendered_path, dest_path, overwrite = TRUE)
      logger::log_success(
        "Report saved as: {dest_path}",
        namespace = "ZooGoN"
      )
    } else {
      logger::log_warn(
        "Expected file not found: {rendered_path}",
        namespace = "ZooGoN"
      )
    }
  }

  invisible(NULL)
}
```

# Darwin Core Archive & GBIF Publication
The function `raw_to_dc()` converts the merged dataset into [**Darwin Core**](https://dwc.tdwg.org/) format following [OBIS](https://obis.org/) and [EMODnet Biology](https://www.emodnet-biology.eu/) standards, producing three extension tables:
- Event: sampling metadata including coordinates (40.81°N, 14.25°E), depth range (0–50 m), and sampling protocol.
- Occurrence: species occurrences linked to WoRMS LSIDs.
- eMoF: quantitative measurements (ind m⁻³) with standardized [BODC NERC](https://vocab.nerc.ac.uk/) vocabulary terms for sex and life stage.
```{r}
#| echo: true
#| code-fold: true
raw_to_dc <- function(
  verbose = TRUE
) {
  conf <- read_config()
  logger::log_info("Starting Darwin Core conversion...", namespace = "ZooGoN")

  logger::log_info("Downloading legacy datasets from hot storage...", namespace = "ZooGoN")
  legacy_84_20 <-
    c("McZoo_84-15.parquet", "McZoo_16-20.parquet") |>
    purrr::map(
      download_sharepoint_file,
      options = conf$storage$sharepoint$credentials,
      bucket = conf$storage$sharepoint$buckets$hot_bucket,
      filename = TRUE
    ) |>
    dplyr::bind_rows()
  logger::log_debug("Legacy data loaded: {nrow(legacy_84_20)} rows", namespace = "ZooGoN")

  logger::log_info("Downloading preprocessed ongoing surveys...", namespace = "ZooGoN")
  ongoing_landings <-
    download_sharepoint_file(
      prefix = conf$ingestion$surveys$preprocessed$file_prefix,
      options = conf$storage$sharepoint$credentials,
      bucket = conf$storage$sharepoint$buckets$automation_bucket,
      format = "parquet"
    )
  logger::log_debug("Ongoing surveys loaded: {nrow(ongoing_landings)} rows", namespace = "ZooGoN")

  merged_data <-
    legacy_84_20 |>
    dplyr::bind_rows(ongoing_landings)
  logger::log_info("Merged data: {nrow(merged_data)} rows", namespace = "ZooGoN")

  # Create Darwin Core Event extension
  logger::log_info("Creating Event extension table...", namespace = "ZooGoN")

  event_ext <- merged_data |>
    dplyr::select(dplyr::all_of(c("eventID", "eventDate"))) |>
    dplyr::arrange(.data$eventDate) |>
    dplyr::distinct() |>
    dplyr::mutate(
      eventDate = as.character(.data$eventDate),
      eventType = "Site Visit",
      decimalLatitude = 40.81,
      decimalLongitude = 14.25,
      geodeticDatum = "WGS84",
      continent = "Europe",
      countryCode = "IT",
      institutionCode = "SZN",
      datasetName = "ZooGoN", # to do: verify dataset name
      locality = "Gulf of Naples",
      stateProvince = "Campania",
      waterBody = "Tyrrhenian Sea",
      maximumDepthInMeters = 50,
      minimumDepthInMeters = 0,
      samplingProtocol = "Vertical tow 0-50m depth"
    )

  # Create Darwin Core Occurrence extension
  logger::log_info("Creating Occurrence extension table...", namespace = "ZooGoN")

  full_table <- merged_data |>
    dplyr::mutate(
      occurrenceStatus = dplyr::if_else(
        .data$individualCount > 0,
        "present",
        "absent"
      ),
      occurrenceID = paste0(.data$eventID, "-occ", dplyr::row_number())
    ) |>
    dplyr::relocate("occurrenceID", .after = "eventID") |>
    dplyr::distinct()

  occurrence_table <- full_table |>
    dplyr::mutate(basisOfRecord = "MachineObservation") |>
    dplyr::select(
      "eventID",
      "occurrenceID",
      "basisOfRecord",
      "scientificName",
      scientificNameID = "lsid",
      "occurrenceStatus"
    )

  # Create Darwin Core eMoF extension
  logger::log_info("Creating eMoF extension table...", namespace = "ZooGoN")

  emof_occ <- build_emof_occurrence(data = full_table)
  emof_events <- build_emof_events(data = full_table)
  emof_table <- dplyr::bind_rows(emof_occ, emof_events)

  # Prepare output
  processing_info <- list(
    processing_date = Sys.time(),
    total_events = nrow(event_ext),
    total_occurrences = nrow(occurrence_table),
    total_measurements = nrow(emof_table),
    date_range = range(event_ext$eventDate, na.rm = TRUE),
    unique_taxa = length(unique(occurrence_table$scientificName))
  )

  metadata_df <- tibble::tibble(
    dataset_title = "40 years of Zooplankton data at LTER MareChiara site (Gulf of Naples, Mediterranean Sea) 1984-2024", # to do: depends on the data we are going to publish
    contact = "Dr. Iole Di Capua (iole.dicapua@szn.it)",
    institution = "Stazione Zoologica Anton Dohrn",
    license = "CC-BY-NC",
    project = "DTO-BioFlow FSTP Grant",
    processing_date = as.character(processing_info$processing_date),
    total_events = processing_info$total_events,
    total_occurrences = processing_info$total_occurrences,
    date_range_start = as.character(processing_info$date_range[1]),
    date_range_end = as.character(processing_info$date_range[2])
  )

  darwin_core_data <- list(
    event = event_ext,
    occurrence = occurrence_table,
    emof = emof_table,
    raw_data = merged_data,
    processing_info = processing_info,
    metadata = metadata_df
  )

  logger::log_info("Events: {processing_info$total_events} | Occurrences: {processing_info$total_occurrences} | Measurements: {processing_info$total_measurements}", namespace = "ZooGoN")
  logger::log_debug("Unique taxa: {processing_info$unique_taxa} | Date range: {processing_info$date_range[1]} to {processing_info$date_range[2]}", namespace = "ZooGoN")

  # Upload to SharePoint
  logger::log_info("Uploading Darwin Core tables to SharePoint...", namespace = "ZooGoN")
  upload_sharepoint_df(
    data = darwin_core_data,
    prefix = conf$ingestion$surveys$darwincore$file_prefix,
    options = conf$storage$sharepoint$credentials,
    bucket = conf$storage$sharepoint$buckets$automation_bucket,
    format = "rds"
  )
  logger::log_success("raw_to_dc complete", namespace = "ZooGoN")
}
```

The function `dc_to_archive()` packages the Darwin Core tables into a Darwin Core Archive (DwC-A) zip file with EML metadata, and uploads it to SharePoint for subsequent publication to [**GBIF**](https://www.gbif.org/).
```{r}
#| echo: true
#| code-fold: true
dc_to_archive <- function() {
  conf <- read_config()

  logger::log_info("Downloading Darwin Core tables from SharePoint...", namespace = "ZooGoN")
  dc_list <-
    download_sharepoint_file(
      prefix = conf$ingestion$surveys$darwincore$file_prefix,
      options = conf$storage$sharepoint$credentials,
      bucket = conf$storage$sharepoint$buckets$automation_bucket,
      format = "rds"
    )
  logger::log_debug("Tables loaded: {nrow(dc_list$event)} events, {nrow(dc_list$occurrence)} occurrences, {nrow(dc_list$emof)} measurements", namespace = "ZooGoN")

  event_df <- dc_list$event

  logger::log_info("Initializing DwC core and extensions...", namespace = "ZooGoN")
  core_event <- LivingNorwayR::initializeGBIFEvent(
    event_df,
    idColumnInfo = "eventID",
    nameAutoMap = TRUE
  )

  occ_ext <- LivingNorwayR::initializeGBIFOccurrence(
    dc_list$occurrence,
    idColumnInfo = "eventID",
    nameAutoMap = TRUE
  )

  emof_ext <- LivingNorwayR::initializeGBIFMeasurementOrFact(
    dc_list$emof,
    idColumnInfo = "eventID",
    nameAutoMap = TRUE
  )

  logger::log_info("Generating EML metadata...", namespace = "ZooGoN")
  eml_obj <- get_metadata(event_df)
  eml_path <- file.path(tempdir(), "mc_eml.xml")

  EML::write_eml(eml_obj, eml_path)
  add_gbif_license_block(eml_path)
  EML::eml_validate(eml_path)
  logger::log_debug("EML validated successfully", namespace = "ZooGoN")

  metadata <- LivingNorwayR::initializeDwCMetadata(
    fileLocation = eml_path,
    fileEncoding = "UTF-8",
    fileType = "eml"
  )

  dwc <- LivingNorwayR::initializeDwCArchive(
    coreDwC = core_event,
    extDwC = list(occ_ext, emof_ext),
    metadata = metadata
  )

  zip_file <- add_version("ZooGoN_dwca", extension = "zip")

  logger::log_info("Exporting DwC-A zip: {basename(zip_file)}", namespace = "ZooGoN")
  dwc$exportAsDwCArchive(
    fileName = zip_file,
    emlLocation = basename(eml_path)
  )

  logger::log_info("Uploading archive to SharePoint...", namespace = "ZooGoN")
  sp_conn <- connect_to_sharepoint(conf$storage$sharepoint$credentials)
  remote_path <- file.path(
    conf$storage$sharepoint$buckets$darwin_core_bucket,
    basename(zip_file)
  )

  upload_file_to_sharepoint(
    file_path = zip_file,
    remote_path = remote_path,
    drive_id = sp_conn$drive_id,
    token = sp_conn$token,
    format = "zip"
  )

  logger::log_success("dc_to_archive complete — uploaded to: {remote_path}", namespace = "ZooGoN")
  invisible(list(archive_path = zip_file, eml_path = eml_path))
}
```

# Example of Darwin Core Output 
```{r}
#| echo: true
dc_list <- ZooGoN::download_sharepoint_file(
  prefix  = conf$ingestion$surveys$darwincore$file_prefix,
  options = conf$storage$sharepoint$credentials,
  bucket  = conf$storage$sharepoint$buckets$automation_bucket,
  format  = "rds"
)

print(head(dc_list$event, 10))
print(head(dc_list$occurrence, 10))
print(head(dc_list$emof, 10))
```

# Report Automation
This report is fully automated and self-regenerating. The pipeline runs every day at 00:00 UTC, ensuring that new field data collected via KoBoToolbox are reflected in the report the following morning. The rendered HTML and PDF outputs are archived as downloadable artifacts on GitHub Actions for 90 days.