---
title: "Gulf of Naples – 40 Years of Zooplankton Biodiversity Assessment"
subtitle: "ZOOGoN-40Y | LTER MareChiara"
author: "Andrea Marcucci (andrea.marcucci@szn.it)"
date: today
format:
  html:
    theme:
      light: flatly
      dark: darkly
    toc: true
    toc-location: right
    toc-depth: 3
    toc-title: "Contents"
    number-sections: false
    code-fold: true
    code-tools: true
    embed-resources: true
    fig-width: 10
    fig-height: 6
    page-layout: article
---

<style>
.taxon-card {
  border-left: 4px solid #2fa4e7;
  padding: 0.75rem 1.25rem;
  border-radius: 4px;
  margin: 1rem 0;
  font-size: 0.95rem;
  background: rgba(47, 164, 231, 0.07);
}
</style>

# Aim

The MareChiara station represents one of the longest-running marine monitoring
sites in the Mediterranean Sea, providing a unique 40-year time series that
enables investigation of long-term zooplankton community dynamics and
variability. This report presents the harmonized taxonomic zooplankton metadata
derived from the LTER dataset, structured in accordance with Darwin Core and
EMODnet Biology standards.

Use the **Taxon Explorer** below to interactively explore the seasonal heatmap
and long-term trend for any taxon in the full dataset.

---

# Data Sources and Pipeline

The three temporal datasets cover 1984–2015, 2016–2020, and 2021–2024, from
the ZOOGoN-40Y project (VLIZ / SZN). Legacy data (1984–2020) are stored in
SharePoint; new surveys since 2021 are collected via
[KoBoToolbox](https://www.kobotoolbox.org). Data are processed through the
[ZooGoN](https://github.com/ioledc/ZOOGoN-40Y) R package — see the
[pipeline vignette](https://ioledc.github.io/ZOOGoN-40Y/articles/data-processing.html)
for details.

```{mermaid}
flowchart LR
A[Build R container] --> B[Ingest surveys]
B --> C[Preprocess surveys]
C --> D[Merge legacy + ongoing data]
D --> E[Build Darwin Core Archive]
D --> F[Render MC survey report]

classDef step fill:#2fa4e7,color:#fff,stroke:#1a8cc0
class A,B,C,D,E,F step
```

---

# Data Loading and Preparation

```{r setup}
#| echo: true
#| code-fold: true
#| message: false
#| warning: false

#run with: R_CONFIG_ACTIVE=production quarto preview inst/report/REPORT_copy.qmd
library(ZooGoN)

# Quarto sets the working directory to inst/report/ — search parent dirs for .env
for (.env_path in c(".env", "../../.env")) {
  if (file.exists(.env_path)) { dotenv::load_dot_env(file = .env_path); break }
}
rm(.env_path)

conf <- ZooGoN::read_config()
pacman::p_load(dplyr, lubridate, worrms, stringr, arrow)

merged <- tryCatch(
  ZooGoN::download_sharepoint_file(
    prefix  = conf$ingestion$tidy_data$file_prefix,
    options = conf$storage$sharepoint$credentials,
    bucket  = conf$storage$sharepoint$buckets$automation_bucket,
    format  = "parquet"
  ),
  error = function(e) {
    warning(sprintf(
      "Could not download tidy data from bucket '%s': %s",
      conf$storage$sharepoint$buckets$automation_bucket,
      conditionMessage(e)
    ))
    NULL
  }
)

# Boolean flag used by downstream chunk eval options (avoids !expr !is.null() YAML conflict)
has_data <- !is.null(merged)
```

```{r data-status}
#| echo: false
#| results: asis

if (is.null(merged)) {
  cat(sprintf(paste0(
    "::: {.callout-warning}\n",
    "## Data not available\n\n",
    "The tidy dataset could not be loaded from SharePoint bucket **`%s`**.\n\n",
    "**To fix this locally:**\n\n",
    "- Set `R_CONFIG_ACTIVE=production` in your R session to read from the ",
    "production bucket, then re-render: `Sys.setenv(R_CONFIG_ACTIVE = \"production\")`\n",
    "- Or run `format_to_tidy()` first to populate the dev bucket.\n\n",
    "The interactive explorer below will be empty until data is available.\n",
    ":::\n"
  ), conf$storage$sharepoint$buckets$automation_bucket))
}
```

```{r process}
#| echo: true
#| code-fold: true
#| message: false
#| warning: false
#| eval: !expr has_data

# --- 1. WoRMS enrichment (Class + Order) with full fallback ---
aphiaid <- merged |>
  dplyr::mutate(aphiaID = as.integer(stringr::str_extract(lsid, "[0-9]+")))

ids <- unique(aphiaid$aphiaID)

get_class_order <- function(id) {
  tryCatch({
    cls <- worrms::wm_classification(id)
    data.frame(
      aphiaID = id,
      class   = if (length(cls$scientificname[cls$rank == "Class"]) > 0)
                  cls$scientificname[cls$rank == "Class"][1] else NA_character_,
      order   = if (length(cls$scientificname[cls$rank == "Order"]) > 0)
                  cls$scientificname[cls$rank == "Order"][1] else NA_character_,
      stringsAsFactors = FALSE
    )
  }, error = function(e) {
    data.frame(aphiaID = id, class = NA_character_, order = NA_character_,
               stringsAsFactors = FALSE)
  })
}

ranks_df <- tryCatch(
  dplyr::bind_rows(lapply(ids, get_class_order)),
  error = function(e) {
    warning(sprintf("WoRMS enrichment failed: %s — taxonomy shown as NA.", conditionMessage(e)))
    data.frame(aphiaID = ids, class = NA_character_, order = NA_character_,
               stringsAsFactors = FALSE)
  }
)

zoo_all <- tryCatch({
  aphiaid |>
    dplyr::left_join(ranks_df, by = "aphiaID") |>
    dplyr::select(-aphiaID) |>
    dplyr::rename(rankClass = class, rankOrder = order)
}, error = function(e) {
  warning(sprintf("WoRMS join failed: %s — proceeding without taxonomy.", conditionMessage(e)))
  merged |> dplyr::mutate(rankClass = NA_character_, rankOrder = NA_character_)
})

# --- 2. Pre-aggregate for OJS ---

# A. Weekly heatmap: one row per taxon × year × week (non-zero only)
heatmap_data <- zoo_all |>
  dplyr::mutate(
    week = lubridate::week(eventDate),
    year = lubridate::year(eventDate)
  ) |>
  dplyr::group_by(year, week, rankClass, rankOrder, scientificName) |>
  dplyr::summarise(abundance = sum(Abundance, na.rm = TRUE), .groups = "drop") |>
  dplyr::filter(abundance > 0) |>
  dplyr::mutate(log_abundance = log(abundance + 1))

# B. Monthly anomaly trend: deseasonalized per-taxon time series
trend_data <- zoo_all |>
  dplyr::group_by(scientificName, rankClass, rankOrder, eventDate) |>
  dplyr::summarise(abundance = sum(Abundance, na.rm = TRUE), .groups = "drop") |>
  dplyr::mutate(
    month      = lubridate::month(eventDate),
    quarter    = as.character(lubridate::quarter(eventDate)),
    year       = lubridate::year(eventDate),
    month_date = as.character(lubridate::floor_date(eventDate, "month")),
    series     = ifelse(year < 1993, "before", "after")
  ) |>
  dplyr::group_by(series, month, scientificName, rankClass, rankOrder) |>
  dplyr::mutate(mean_abund = mean(abundance), anomaly = abundance - mean_abund) |>
  dplyr::group_by(series, month_date, quarter, scientificName, rankClass, rankOrder) |>
  dplyr::summarise(anomaly = mean(anomaly), .groups = "drop")

# C. Taxon metadata for the selector
taxa_meta <- zoo_all |>
  dplyr::distinct(scientificName, rankClass, rankOrder) |>
  dplyr::arrange(scientificName)

ojs_define(heatmap_data = heatmap_data)
ojs_define(trend_data   = trend_data)
ojs_define(taxa_meta    = taxa_meta)
```

```{r ojs-fallback}
#| echo: false
#| include: false
#| eval: !expr isFALSE(has_data)

# When data loading failed, pass empty frames so OJS cells still render
ojs_define(heatmap_data = data.frame(
  scientificName = character(), rankClass = character(), rankOrder = character(),
  year = integer(), week = integer(), abundance = numeric(), log_abundance = numeric()
))
ojs_define(trend_data = data.frame(
  scientificName = character(), rankClass = character(), rankOrder = character(),
  month_date = character(), quarter = character(), series = character(),
  anomaly = numeric()
))
ojs_define(taxa_meta = data.frame(
  scientificName = character(), rankClass = character(), rankOrder = character()
))
```

---

# Taxon Explorer

::: {.callout-note collapse="true"}
## Note: interactive features require HTTP

OJS plots work when this document is served via HTTP. Open via
`quarto preview` locally, or view the file through the
[GitHub Actions artifact viewer](https://github.com/ioledc/ZOOGoN-40Y/actions)
(not via `file://`).
:::

```{ojs}
//| echo: false

heatmap_js = transpose(heatmap_data)
trend_js   = transpose(trend_data)
taxa_js    = transpose(taxa_meta)
```

Type in **Filter taxa** to narrow the list by species name, class or order,
then choose from the **Select taxon** dropdown. Both charts update instantly.

```{ojs}
//| echo: false

viewof search_query = Inputs.text({
  label: "Filter taxa",
  placeholder: "Type a species name, class or order…",
  width: 420
})
```

```{ojs}
//| echo: false

filtered_names = (() => {
  const q = search_query.toLowerCase()
  const names = taxa_js
    .filter(d =>
      q === "" ||
      d.scientificName.toLowerCase().includes(q) ||
      (d.rankClass && d.rankClass.toLowerCase().includes(q)) ||
      (d.rankOrder && d.rankOrder.toLowerCase().includes(q))
    )
    .map(d => d.scientificName)
    .sort()
  return names.length > 0 ? names : taxa_js.map(d => d.scientificName).sort()
})()

viewof selected = Inputs.select(filtered_names, {
  label: "Select taxon",
  width: 420
})
```

```{ojs}
//| echo: false

// Summary card — updates reactively when selected changes
{
  const rows     = heatmap_js.filter(d => d.scientificName === selected)
  const total    = d3.sum(rows, d => d.abundance)
  const years    = rows.map(d => +d.year)
  const n_years  = new Set(years).size
  const min_year = d3.min(years)
  const max_year = d3.max(years)
  const meta     = taxa_js.find(d => d.scientificName === selected) || {}
  return htl.html`<div class="taxon-card">
    <strong style="font-size:1.05em"><em>${selected}</em></strong><br>
    Class: <strong>${meta.rankClass ?? "—"}</strong> &ensp;|&ensp;
    Order: <strong>${meta.rankOrder ?? "—"}</strong> &ensp;|&ensp;
    Observed: <strong>${min_year ?? "—"}–${max_year ?? "—"}</strong>
    (${n_years} yr) &ensp;|&ensp;
    Total: <strong>${d3.format(",.0f")(total)}</strong> ind/m³
  </div>`
}
```

::: {.panel-tabset}

## Seasonal Heatmap

Each cell shows total abundance (ind/m³) for a given **week** (x-axis) and
**year** (y-axis), log-transformed (log(x + 1)) to compress the dynamic range.
Missing cells indicate no sampling or zero abundance for that period. Hover over
a cell for the exact values.

```{ojs}
//| echo: false

{
  const data = heatmap_js
    .filter(d => d.scientificName === selected)
    .map(d => ({ ...d, week: +d.week, year: +d.year }))

  return Plot.plot({
    height: 540,
    marginLeft: 52,
    marginRight: 20,
    style: { fontSize: "13px" },
    color: {
      scheme: "turbo",
      legend: true,
      label: "log(ind/m³ + 1)",
      width: 280
    },
    x: {
      type: "band",
      label: "Week of year",
      tickValues: [1, 5, 10, 15, 20, 25, 30, 35, 40, 45, 50]
    },
    y: {
      type: "band",
      label: "Year"
    },
    marks: [
      Plot.cell(data, {
        x: d => String(d.week),
        y: d => String(d.year),
        fill: "log_abundance",
        tip: true,
        inset: 0.5
      })
    ]
  })
}
```

## Long-term Trend

Deseasonalized monthly abundance anomaly. Points show deviation from the
long-term monthly mean, coloured by quarter (Q1–Q4). Linear regression lines
are fitted to the post-1993 period (after the sampling gap) by quarter.

```{ojs}
//| echo: false

{
  const pts   = trend_js
    .filter(d => d.scientificName === selected)
    .map(d => ({ ...d, date: new Date(d.month_date) }))
    .sort((a, b) => a.date - b.date)
  const after = pts.filter(d => d.series === "after")

  return Plot.plot({
    height: 420,
    marginLeft: 60,
    marginRight: 20,
    style: { fontSize: "13px" },
    color: {
      legend: true,
      label: "Quarter",
      domain: ["1", "2", "3", "4"],
      scheme: "Spectral"
    },
    x: { type: "utc", label: "Date" },
    y: { label: "Anomaly (ind/m³)", zero: true },
    marks: [
      Plot.ruleY([0], { stroke: "#aaa", strokeDasharray: "4 4" }),
      Plot.line(pts, {
        x: "date",
        y: "anomaly",
        stroke: "#bbb",
        strokeOpacity: 0.25,
        strokeWidth: 0.8
      }),
      Plot.dot(pts, {
        x: "date",
        y: "anomaly",
        fill: "quarter",
        fillOpacity: 0.65,
        r: 3,
        tip: true
      }),
      Plot.linearRegressionY(after, {
        x: "date",
        y: "anomaly",
        stroke: "quarter",
        strokeWidth: 2.5
      })
    ]
  })
}
```

:::

---

# Dataset Overview

All `r if (exists("taxa_meta")) nrow(taxa_meta) else "–"` taxa in the 40-year
dataset. Click a column header to sort. Use the search box in the Taxon Explorer
to narrow the selection.

```{ojs}
//| echo: false

Inputs.table(taxa_js, {
  columns: ["scientificName", "rankClass", "rankOrder"],
  header: {
    scientificName: "Scientific Name",
    rankClass: "Class",
    rankOrder: "Order"
  },
  sort: "scientificName",
  width: { scientificName: 320, rankClass: 190, rankOrder: 190 }
})
```

---

# Automated Data Pipeline

This report is produced by a fully automated pipeline running daily at 00:00 UTC
via [GitHub Actions](https://github.com/ioledc/ZOOGoN-40Y/actions). HTML
artifacts are retained for 90 days and are accessible directly through the
Actions UI. All survey data are simultaneously converted into Darwin Core–compliant
format for integration with OBIS, EMODnet Biology, and GBIF.
