---
title: "Gulf of Naples – 40 Years of Zooplankton Biodiversity Assessment"
subtitle: "ZOOGoN-40Y | LTER MareChiara"
author: "Andrea Marcucci (andrea.marcucci@szn.it) & Lorenzo Longobardi (lorenzo.longobardi@gmail.com)"
date: today
format:
  html:
    theme:
      light: flatly
      #dark: darkly
    toc: true
    toc-location: right
    toc-depth: 3
    toc-title: "Contents"
    number-sections: false
    code-fold: true
    code-tools: true
    embed-resources: true
    fig-width: 10
    fig-height: 6
    page-layout: article
---

<style>
#zg-heatmap { height: 720px !important; }
#zg-trend { height: 520px !important; }
#zg-phenology { height: 480px !important; }

.taxon-card {
  border-left: 4px solid #1f78b4;
  padding: 0.75rem 1.25rem;
  border-radius: 6px;
  margin: 1rem 0;
  font-size: 0.95rem;
  background: rgba(31, 120, 180, 0.07);
}
.section-intro {
  color: #555;
  font-size: 0.97rem;
  line-height: 1.6;
  margin-bottom: 0.75rem;
}
</style>

# Aim

The MareChiara station represents one of the longest-running marine monitoring
sites in the Mediterranean Sea, providing a unique 40-year time series that
enables investigation of long-term zooplankton community dynamics and
variability. This report presents the harmonized taxonomic zooplankton metadata
derived from the LTER dataset, structured in accordance with Darwin Core and
EMODnet Biology standards.

Use the **Taxon Explorer** below to interactively explore the seasonal heatmap
and long-term trend for any taxon in the full dataset.

---

# Data Sources and Pipeline

The three temporal datasets cover 1984–2015, 2016–2020, and 2021–2024, from
the ZOOGoN-40Y project (VLIZ / SZN). Legacy data (1984–2020) are stored in
SharePoint; new surveys since 2021 are collected via
[KoBoToolbox](https://www.kobotoolbox.org). Data are processed through the
[ZooGoN](https://github.com/ioledc/ZOOGoN-40Y) R package — see the
[pipeline vignette](https://ioledc.github.io/ZOOGoN-40Y/articles/data-processing.html)
for details.

```{mermaid}
flowchart LR
A[Build R container] --> B[Ingest surveys]
B --> C[Preprocess surveys]
C --> D[Merge legacy + ongoing data]
D --> E[Build Darwin Core Archive]
D --> F[Render MC survey report]

classDef step fill:#2fa4e7,color:#fff,stroke:#1a8cc0
class A,B,C,D,E,F step
```

---

# Data Loading and Preparation

```{r setup}
#| echo: true
#| code-fold: true
#| message: false
#| warning: false

#run with: R_CONFIG_ACTIVE=production quarto preview inst/report/REPORT_copy.qmd
library(ZooGoN)

# Quarto sets the working directory to inst/report/ — search parent dirs for .env
for (.env_path in c(".env", "../../.env")) {
  if (file.exists(.env_path)) { dotenv::load_dot_env(file = .env_path); break }
}
rm(.env_path)

conf <- ZooGoN::read_config()
pacman::p_load(dplyr, lubridate, worrms, stringr, arrow,
               highcharter, DT, jsonlite, RColorBrewer, htmltools, htmlwidgets)

merged <- tryCatch(
  ZooGoN::download_sharepoint_file(
    prefix  = conf$ingestion$tidy_data$file_prefix,
    options = conf$storage$sharepoint$credentials,
    bucket  = conf$storage$sharepoint$buckets$automation_bucket,
    format  = "parquet"
  ),
  error = function(e) {
    warning(sprintf(
      "Could not download tidy data from bucket '%s': %s",
      conf$storage$sharepoint$buckets$automation_bucket,
      conditionMessage(e)
    ))
    NULL
  }
)

# Boolean flag used by downstream chunk eval options (avoids !expr !is.null() YAML conflict)
has_data <- !is.null(merged)
```

```{r data-status}
#| echo: false
#| results: asis

if (is.null(merged)) {
  cat(sprintf(paste0(
    "::: {.callout-warning}\n",
    "## Data not available\n\n",
    "The tidy dataset could not be loaded from SharePoint bucket **`%s`**.\n\n",
    "**To fix this locally:**\n\n",
    "- Set `R_CONFIG_ACTIVE=production` in your R session to read from the ",
    "production bucket, then re-render: `Sys.setenv(R_CONFIG_ACTIVE = \"production\")`\n",
    "- Or run `format_to_tidy()` first to populate the dev bucket.\n\n",
    "The interactive explorer below will be empty until data is available.\n",
    ":::\n"
  ), conf$storage$sharepoint$buckets$automation_bucket))
}
```

```{r process}
#| echo: true
#| code-fold: true
#| message: false
#| warning: false
#| eval: !expr has_data

# --- 1. WoRMS enrichment (Class + Order) with full fallback ---
aphiaid <- merged |>
  dplyr::mutate(aphiaID = as.integer(stringr::str_extract(lsid, "[0-9]+")))

ids <- na.omit(unique(aphiaid$aphiaID))

get_class_order <- function(id) {
  Sys.sleep(0.1) # Respect WoRMS API rate limits
  tryCatch({
    cls <- worrms::wm_classification(id)
    data.frame(
      aphiaID = id,
      class   = if (length(cls$scientificname[cls$rank == "Class"]) > 0)
                  cls$scientificname[cls$rank == "Class"][1] else NA_character_,
      order   = if (length(cls$scientificname[cls$rank == "Order"]) > 0)
                  cls$scientificname[cls$rank == "Order"][1] else NA_character_,
      stringsAsFactors = FALSE
    )
  }, error = function(e) {
    data.frame(aphiaID = id, class = NA_character_, order = NA_character_,
               stringsAsFactors = FALSE)
  })
}

ranks_df <- tryCatch(
  dplyr::bind_rows(lapply(ids, get_class_order)),
  error = function(e) {
    warning(sprintf("WoRMS enrichment failed: %s — taxonomy shown as NA.", conditionMessage(e)))
    data.frame(aphiaID = ids, class = NA_character_, order = NA_character_,
               stringsAsFactors = FALSE)
  }
)

zoo_all <- tryCatch({
  aphiaid |>
    dplyr::left_join(ranks_df, by = "aphiaID") |>
    dplyr::select(-aphiaID) |>
    dplyr::rename(rankClass = class, rankOrder = order)
}, error = function(e) {
  warning(sprintf("WoRMS join failed: %s — proceeding without taxonomy.", conditionMessage(e)))
  merged |> dplyr::mutate(rankClass = NA_character_, rankOrder = NA_character_)
})

# --- 2. Pre-aggregate data ---

# A. Weekly heatmap: one row per taxon × year × week (non-zero only)
heatmap_data <- zoo_all |>
  dplyr::mutate(
    week = lubridate::week(eventDate),
    year = lubridate::year(eventDate)
  ) |>
  dplyr::group_by(year, week, rankClass, rankOrder, scientificName) |>
  dplyr::summarise(abundance = sum(Abundance, na.rm = TRUE), .groups = "drop") |>
  dplyr::filter(abundance > 0) |>
  dplyr::mutate(log_abundance = log(abundance + 1))

# B. Monthly anomaly trend: deseasonalized per-taxon time series
trend_data <- zoo_all |>
  dplyr::group_by(scientificName, rankClass, rankOrder, eventDate) |>
  dplyr::summarise(abundance = sum(Abundance, na.rm = TRUE), .groups = "drop") |>
  dplyr::mutate(
    month      = lubridate::month(eventDate),
    quarter    = as.character(lubridate::quarter(eventDate)),
    year       = lubridate::year(eventDate),
    month_date = as.character(lubridate::floor_date(eventDate, "month")),
    series     = ifelse(year < 1993, "before", "after")
  ) |>
  dplyr::group_by(series, month, scientificName, rankClass, rankOrder) |>
  dplyr::mutate(mean_abund = mean(abundance), anomaly = abundance - mean_abund) |>
  dplyr::group_by(series, month_date, quarter, scientificName, rankClass, rankOrder) |>
  dplyr::summarise(anomaly = mean(anomaly), .groups = "drop")

# C. Taxon metadata for the selector
taxa_meta <- zoo_all |>
  dplyr::distinct(scientificName, rankClass, rankOrder, lsid) |>
  dplyr::mutate(
    id = stringr::str_extract(lsid, "[0-9]+"),
    link = paste0(
      "https://www.marinespecies.org/aphia.php?p=taxdetails&id=",
      id
    )
  ) |>
  dplyr::select(-c("id", "lsid")) |>
  dplyr::relocate(link, .after = "scientificName") |>
  dplyr::arrange(scientificName)

# D. Annual taxonomic composition by Class (community-level)
composition_data <- zoo_all |>
  dplyr::mutate(year = lubridate::year(eventDate)) |>
  dplyr::group_by(year, rankClass) |>
  dplyr::summarise(abundance = sum(Abundance, na.rm = TRUE), .groups = "drop") |>
  dplyr::group_by(year) |>
  dplyr::mutate(total = sum(abundance), pct = abundance / total * 100) |>
  dplyr::ungroup() |>
  dplyr::filter(!is.na(rankClass))

# E. Annual peak week per taxon (phenology)
phenology_data <- zoo_all |>
  dplyr::mutate(
    week = lubridate::week(eventDate),
    year = lubridate::year(eventDate)
  ) |>
  dplyr::group_by(scientificName, rankClass, rankOrder, year, week) |>
  dplyr::summarise(abundance = sum(Abundance, na.rm = TRUE), .groups = "drop") |>
  dplyr::group_by(scientificName, rankClass, rankOrder, year) |>
  dplyr::slice_max(abundance, n = 1, with_ties = FALSE) |>
  dplyr::ungroup() |>
  dplyr::rename(peak_week = week, peak_abundance = abundance)

# F. Regression line endpoints for trend chart (per taxon × quarter, post-1993)
trend_regression <- trend_data |>
  dplyr::filter(series == "after") |>
  dplyr::mutate(date_num = as.numeric(as.Date(month_date))) |>
  dplyr::group_by(scientificName, quarter) |>
  dplyr::filter(dplyr::n() >= 3) |>
  dplyr::group_modify(~ {
    m  <- lm(anomaly ~ date_num, data = .x)
    x0 <- min(.x$date_num); x1 <- max(.x$date_num)
    dplyr::tibble(
      x_min_ms = x0 * 86400000, x_max_ms = x1 * 86400000,
      y_min    = as.numeric(predict(m, data.frame(date_num = x0))),
      y_max    = as.numeric(predict(m, data.frame(date_num = x1)))
    )
  }, .keep = FALSE) |>
  dplyr::ungroup()

# G. Regression line endpoints for phenology chart (per taxon)
phenology_regression <- phenology_data |>
  dplyr::group_by(scientificName) |>
  dplyr::filter(dplyr::n() >= 3) |>
  dplyr::group_modify(~ {
    m  <- lm(peak_week ~ year, data = .x)
    x0 <- min(.x$year); x1 <- max(.x$year)
    dplyr::tibble(
      x_min = x0, x_max = x1,
      y_min = as.numeric(predict(m, data.frame(year = x0))),
      y_max = as.numeric(predict(m, data.frame(year = x1)))
    )
  }, .keep = FALSE) |>
  dplyr::ungroup()

# H. Serialise everything for client-side JS (file:// compatible, no OJS)
all_years_rev <- seq(
  max(2024, max(lubridate::year(zoo_all$eventDate), na.rm = TRUE)),
  1984
)
year_idx_map  <- setNames(seq_along(all_years_rev) - 1L, as.character(all_years_rev))

zoogon_json <- jsonlite::toJSON(list(
  yearCats  = as.character(all_years_rev),
  yearIdx   = as.list(year_idx_map),
  heatmap   = heatmap_data   |> dplyr::select(scientificName, week, year, log_abundance, abundance),
  trend     = trend_data |>
    dplyr::mutate(month_ms = as.numeric(as.POSIXct(month_date, tz = "UTC")) * 1000) |>
    dplyr::select(scientificName, month_ms, quarter, anomaly),
  trendReg  = trend_regression,
  phenology = phenology_data |> dplyr::select(scientificName, year, peak_week, peak_abundance),
  phenoReg  = phenology_regression,
  meta      = taxa_meta
), auto_unbox = TRUE, na = "null", digits = 4)

first_taxon <- taxa_meta |>
  dplyr::filter(scientificName %in% unique(heatmap_data$scientificName)) |>
  dplyr::pull(scientificName) |>
  as.character() |>
  head(1)

if (length(first_taxon) == 0 || is.na(first_taxon)) {
  first_taxon <- if (nrow(taxa_meta) > 0) as.character(taxa_meta$scientificName[1]) else ""
}
```

```{r data-fallback}
#| echo: false
#| include: false
#| eval: !expr isFALSE(has_data)

# When data loading failed, initialise empty objects so downstream chunks
# can still reference these variables without error
heatmap_data        <- data.frame(scientificName=character(), week=integer(),
                                   year=integer(), log_abundance=numeric(), abundance=numeric())
trend_data          <- data.frame(scientificName=character(), month_date=character(),
                                   quarter=character(), series=character(), anomaly=numeric())
taxa_meta           <- data.frame(scientificName=character(), rankClass=character(), rankOrder=character())
composition_data    <- data.frame(year=integer(), rankClass=character(), pct=numeric())
phenology_data      <- data.frame(scientificName=character(), year=integer(),
                                   peak_week=integer(), peak_abundance=numeric())
trend_regression    <- data.frame(scientificName=character(), quarter=character(),
                                   x_min_ms=numeric(), x_max_ms=numeric(),
                                   y_min=numeric(), y_max=numeric())
phenology_regression <- data.frame(scientificName=character(),
                                    x_min=integer(), x_max=integer(),
                                    y_min=numeric(), y_max=numeric())
all_years_rev <- integer(0)
year_idx_map  <- setNames(integer(0), character(0))
zoogon_json   <- "{}"
first_taxon   <- ""
```

---

# Community Overview

Annual relative abundance (%) of each taxonomic Class across all samples.
The stacked area chart reveals shifts in community composition over the 40-year
record. Hover over any area for exact values.

```{r composition-chart}
#| echo: false
#| message: false
if (!has_data) {
  htmltools::p("Data not available.", style = "color:#888; font-style:italic")
} else {
  classes <- composition_data |>
    dplyr::group_by(rankClass) |>
    dplyr::summarise(tot = sum(pct, na.rm = TRUE), .groups = "drop") |>
    dplyr::arrange(dplyr::desc(tot)) |>
    dplyr::pull(rankClass)
  years_v <- sort(unique(composition_data$year))

  hc <- highchart() |>
    hc_chart(type = "area", animation = FALSE) |>
    hc_xAxis(categories = as.character(years_v), tickInterval = 5,
             title = list(text = "Year")) |>
    hc_yAxis(title = list(text = "Relative abundance (%)"), max = 100, min = 0) |>
    hc_plotOptions(area = list(stacking = "normal", lineWidth = 0.5,
                               marker = list(enabled = FALSE))) |>
    hc_tooltip(shared = TRUE, valueDecimals = 1, valueSuffix = "%") |>
    hc_legend(enabled = TRUE)

  for (cls in classes) {
    vals <- composition_data |>
      dplyr::filter(rankClass == cls) |>
      dplyr::right_join(data.frame(year = years_v), by = "year") |>
      dplyr::arrange(year) |>
      dplyr::pull(pct)
    hc <- hc |>
      hc_add_series(name = cls, data = as.list(replace(vals, is.na(vals), 0)), type = "area")
  }
  hc
}
```

---

# Taxon Explorer

Type in **Filter taxa** to narrow the list by species name, class or order,
then choose from the **Select taxon** dropdown. All charts update instantly.

```{r explorer-data}
#| echo: false
# type="application/json" is never executed by the browser and immune to
# </script> injection — the only safe way to embed large JSON blobs in HTML
htmltools::tags$script(
  type = "application/json",
  id   = "zoogon-data",
  htmltools::HTML(zoogon_json)
)
```

```{r explorer-ui}
#| echo: false
# Pure HTML — no inline event attributes; JS listeners added by zoogonInit() in explorer-js
htmltools::tags$div(
  style = "display:flex; gap:1.5rem; align-items:flex-end; margin:1rem 0; flex-wrap:wrap",
  htmltools::tags$div(
    htmltools::tags$label("Filter taxa", `for` = "taxon-search",
      style = "display:block; font-size:.875rem; margin-bottom:.25rem; font-weight:500"),
    htmltools::tags$input(id = "taxon-search", type = "text",
      placeholder = "Type species, class or order\u2026",
      style = "width:300px; padding:.35rem .5rem; border:1px solid #ccc;
               border-radius:4px; font-size:.9rem")
  ),
  htmltools::tags$div(
    htmltools::tags$label("Select taxon", `for` = "taxon-select",
      style = "display:block; font-size:.875rem; margin-bottom:.25rem; font-weight:500"),
    htmltools::tags$select(id = "taxon-select",
      style = "width:300px; padding:.35rem .5rem; border:1px solid #ccc;
               border-radius:4px; font-size:.9rem")
  )
)
```

```{r explorer-card}
#| echo: false
htmltools::div(id = "taxon-card", class = "taxon-card",
  htmltools::div(id = "taxon-card-content", "\u2014"))
```

::: {.panel-tabset}

## Seasonal Heatmap

Each cell shows total abundance (ind/m³) for a given **week** (x-axis) and
**year** (y-axis). Missing cells indicate no sampling or zero abundance for that
period. Hover over a cell for the exact value.

```{r heatmap-chart}
#| echo: false
#| message: false
if (!has_data) {
  htmltools::p("Data not available.", style = "color:#888; font-style:italic")
} else {
  yl_colors <- color_stops(9, RColorBrewer::brewer.pal(9, "YlGnBu"))

  hm_init <- heatmap_data |> dplyr::filter(scientificName == first_taxon) |>
    dplyr::mutate(
      x     = week - 1L,
      y     = as.integer(year_idx_map[as.character(year)]),
      value = log_abundance
    ) |>
    dplyr::select(x, y, value)

  highchart(height = 720) |>
    hc_chart(type = "heatmap", animation = FALSE, marginBottom = 80) |>
    hc_add_series(
      data = highcharter::list_parse2(hm_init),
      name = "ind/m\u00b3",
      type = "heatmap"
    ) |>
    hc_xAxis(
      categories = as.character(1:52), 
      min = 0, max = 51,
      tickInterval = 4,
      title = list(text = "Week of year")
    ) |>
    hc_yAxis(categories = as.character(all_years_rev), 
             min = 0, max = length(all_years_rev) - 1,
             tickInterval = 4, title = list(text = "Year"), reversed = TRUE) |>
    hc_colorAxis(stops = yl_colors, min = 0, nullColor = "white") |>
    hc_tooltip(formatter = JS(
      "function(){
         return '<b>Week ' + this.series.xAxis.categories[this.point.x] + '</b>' +
                ' \u00d7 <b>' + this.series.yAxis.categories[this.point.y] + '</b><br>' +
                'ind/m\u00b3: ' + Highcharts.numberFormat(Math.round(Math.exp(this.point.value) - 1), 0);
       }"
    )) |>
    hc_legend(layout = "horizontal", align = "center", verticalAlign = "top") |>
    hc_size(height = 720) |>
    htmlwidgets::onRender(
      "function(el,x){
         el.id = 'zg-heatmap';
         var g = window.__zoogon__ = window.__zoogon__||{};
         g.charts = g.charts||{};
         g.charts.heatmap = Highcharts.charts.find(function(c){ return c && c.renderTo === el; });
         if (g.meta && g.meta.length) {
           var sel = document.getElementById('taxon-select');
           zoogonUpdate(sel && sel.value ? sel.value : g.meta[0].scientificName);
         }
       }"
    )
}
```

## Long-term Trend

Deseasonalized monthly abundance anomaly. Points show deviation from the
long-term monthly mean, coloured by quarter (Q1–Q4). Linear regression lines
are fitted to the post-1993 period (after the sampling gap) by quarter.

```{r trend-chart}
#| echo: false
#| message: false
if (!has_data) {
  htmltools::p("Data not available.", style = "color:#888; font-style:italic")
} else {
  qcols    <- c("1" = "#e41a1c", "2" = "#377eb8", "3" = "#4daf4a", "4" = "#984ea3")
  tr_init  <- trend_data        |> dplyr::filter(scientificName == first_taxon)
  re_init  <- trend_regression  |> dplyr::filter(scientificName == first_taxon)

 hc_tr <- highchart(height = 520) |>
  hc_chart(animation = FALSE) |>
  hc_xAxis(type = "datetime", title = list(text = "Date")) |>
  hc_yAxis(
    title = list(text = "Anomaly (ind/m\u00b3)"),
    plotLines = list(list(value = 0, color = "#888", dashStyle = "Dash", width = 1))
  )

for (q in c("1", "2", "3", "4")) {
  pts <- tr_init |>
    dplyr::filter(quarter == q) |>
    dplyr::arrange(month_date)
  
  if (nrow(pts) > 0) {
    pts <- pts |>
      dplyr::mutate(x = as.numeric(as.POSIXct(month_date, tz = "UTC")) * 1000, y = anomaly) |>
      dplyr::select(x, y)
  } else {
    pts <- data.frame(x = numeric(0), y = numeric(0))
  }
    
  hc_tr <- hc_tr |>
    hc_add_series(
      data   = highcharter::list_parse2(pts),
      type   = "scatter", 
      name   = paste0("Q", q), 
      color  = qcols[[as.character(q)]], 
      marker = list(radius = 3.5, symbol = "circle",
                    lineWidth = 0.5, lineColor = "white")
    )
}

for (q in c("1", "2", "3", "4")) {
  r  <- re_init |> dplyr::filter(quarter == q)
  
  if (nrow(r) > 0) { 
    rd <- data.frame(x = c(r$x_min_ms, r$x_max_ms), y = c(r$y_min, r$y_max))
  } else {
    rd <- data.frame(x = numeric(0), y = numeric(0))
  }
    
  hc_tr <- hc_tr |>
    hc_add_series(
      data = highcharter::list_parse2(rd), type = "line",
      name = paste0("Trend Q", q), 
      color = qcols[[as.character(q)]],
      lineWidth = 2.5, showInLegend = FALSE,
      marker = list(enabled = FALSE), enableMouseTracking = FALSE
    )
}

hc_tr |>
  hc_size(height = 520) |>
  htmlwidgets::onRender(
    "function(el,x){
       var g = window.__zoogon__ = window.__zoogon__||{};
       g.charts = g.charts||{};
       
       el.id = 'zg-trend'; 
       
       g.charts.trend = Highcharts.charts.find(function(c){ return c && c.renderTo === el; });
       if (g.meta && g.meta.length) {
         var sel = document.getElementById('taxon-select');
         // Check if zoogonUpdate exists before calling to prevent JS crashes
         if (typeof zoogonUpdate === 'function') {
           zoogonUpdate(sel && sel.value ? sel.value : g.meta[0].scientificName);
         }
       }
     }"
  )
}
```

## Phenology Shift

Week of peak abundance per year, showing whether the seasonal timing of the
selected taxon has shifted over the 40-year record. The dashed line marks
mid-year (week 26). The regression line indicates the long-term phenological
trend.

```{r phenology-chart}
#| echo: false
#| message: false
if (!has_data) {
  htmltools::p("Data not available.", style = "color:#888; font-style:italic")
} else {
  ph_init  <- phenology_data       |> dplyr::filter(scientificName == first_taxon) |> dplyr::arrange(year)
  pr_init  <- phenology_regression |> dplyr::filter(scientificName == first_taxon)
  pr_line  <- if (nrow(pr_init) > 0)
    data.frame(x = c(pr_init$x_min, pr_init$x_max), y = c(pr_init$y_min, pr_init$y_max))
  else
    data.frame(x = numeric(0), y = numeric(0))

  highchart(height = 480) |>
    hc_chart(animation = FALSE) |>
    hc_add_series(
      data   = highcharter::list_parse2(
        ph_init |> dplyr::mutate(x = year, y = peak_week) |> dplyr::select(x, y)
      ),
      type   = "scatter", name = "Peak week", color = "#2166ac",
      marker = list(radius = 5, symbol = "circle")
    ) |>
    hc_add_series(
      data = highcharter::list_parse2(pr_line), type = "line",
      name = "Trend", color = "#d73027", lineWidth = 2.2,
      marker = list(enabled = FALSE), showInLegend = FALSE,
      enableMouseTracking = FALSE
    ) |>
    hc_xAxis(title = list(text = "Year")) |>
    hc_yAxis(
      title = list(text = "Peak week of year"), min = 1, max = 52,
      plotLines = list(list(value = 26, color = "#aaa", dashStyle = "Dash", width = 1))
    ) |>
    hc_size(height = 480) |>
    htmlwidgets::onRender(
      "function(el,x){
         var g = window.__zoogon__ = window.__zoogon__||{};
         g.charts = g.charts||{};
         el.id = 'zg-phenology';
         g.charts.phenology = Highcharts.charts.find(function(c){ return c && c.renderTo === el; });
         if (g.meta && g.meta.length) {
           var sel = document.getElementById('taxon-select');
           zoogonUpdate(sel && sel.value ? sel.value : g.meta[0].scientificName);
         }
       }"
    )
}
```

:::

```{r explorer-js}
#| echo: false
htmltools::tags$script(htmltools::HTML('
function zoogonFilter() {
  var g = window.__zoogon__;
  if (!g || !g.meta) return;
  var q   = document.getElementById("taxon-search").value.toLowerCase();
  var sel = document.getElementById("taxon-select");
  var cur = sel.value;
  var flt = q === "" ? g.meta : g.meta.filter(function(t) {
    return t.scientificName.toLowerCase().includes(q) ||
           (t.rankClass && t.rankClass.toLowerCase().includes(q)) ||
           (t.rankOrder && t.rankOrder.toLowerCase().includes(q));
  });
  if (!flt.length) flt = g.meta;
  sel.innerHTML = "";
  var kept = false;
  flt.forEach(function(t) {
    var o = document.createElement("option");
    o.value = t.scientificName; o.textContent = t.scientificName;
    if (t.scientificName === cur) { o.selected = true; kept = true; }
    sel.appendChild(o);
  });
  zoogonUpdate(kept ? cur : flt[0].scientificName);
}

function getChart(elId) {
  var el = document.getElementById(elId);
  if (!el) return null;
  return Highcharts.charts.find(function(c){ return c && c.renderTo === el; }) || null;
}

function zoogonUpdate(taxon) {
  var g = window.__zoogon__;
  if (!g) return;
  var hm   = g.heatmap.filter(function(d) { return d.scientificName === taxon; });
  var tr   = g.trend.filter(function(d)   { return d.scientificName === taxon; });
  var ph   = g.phenology.filter(function(d){ return d.scientificName === taxon; });
  var meta = g.meta.find(function(d) { return d.scientificName === taxon; }) || {};

  /* ---- Summary card ---- */
  var years = hm.map(function(d) { return d.year; });
  var total = hm.reduce(function(s, d) { return s + d.abundance; }, 0);
  var minY  = years.length ? Math.min.apply(null, years) : "\u2014";
  var maxY  = years.length ? Math.max.apply(null, years) : "\u2014";
  document.getElementById("taxon-card-content").innerHTML =
    "<em>" + taxon + "</em><br>" +
    "Class: <strong>" + (meta.rankClass || "\u2014") + "</strong> &ensp;|&ensp;" +
    "Order: <strong>" + (meta.rankOrder || "\u2014") + "</strong> &ensp;|&ensp;" +
    "Observed: <strong>" + minY + "\u2013" + maxY + "</strong>" +
    " (" + new Set(years).size + " yr) &ensp;|&ensp;" +
    "Total: <strong>" + total.toLocaleString() + "</strong> ind/m\u00b3";

  /* ---- Heatmap (full year × week grid, null for missing cells) ---- */
  var hmChart = (g.charts && g.charts.heatmap) || getChart("zg-heatmap");
  if (hmChart) { g.charts = g.charts||{}; g.charts.heatmap = hmChart; }
  if (hmChart) {
    var hmMap = {};
    hm.forEach(function(d) {
      hmMap[(d.week - 1) + "_" + g.yearIdx[String(d.year)]] = d.log_abundance;
    });
    var nYears = g.yearCats.length;
    var hmData = [];
    for (var yi = 0; yi < nYears; yi++) {
      for (var w = 0; w < 52; w++) {
        var v = hmMap[w + "_" + yi];
        if (v !== undefined && v !== null) {
          hmData.push([w, yi, v]);
        }
      }
    }
    hmChart.series[0].setData(hmData, true, { duration: 200 });
  }

  /* ---- Trend: 4 scatter series + 4 regression lines ---- */
  var trChart = (g.charts && g.charts.trend) || getChart("zg-trend");
  if (trChart) { g.charts = g.charts||{}; g.charts.trend = trChart; }
  if (trChart) {
    ["1","2","3","4"].forEach(function(q, i) {
      var pts = tr.filter(function(d) { return String(d.quarter) === q; })
        .map(function(d) { return [d.month_ms, d.anomaly]; })
        .sort(function(a, b) { return a[0] - b[0]; });
      trChart.series[i].setData(pts, false);
    });
    var regs = g.trendReg || [];
    ["1","2","3","4"].forEach(function(q, i) {
      var r  = regs.find(function(d) { return d.scientificName === taxon && String(d.quarter) === q; });
      var rd = r ? [[r.x_min_ms, r.y_min], [r.x_max_ms, r.y_max]] : [];
      trChart.series[4 + i].setData(rd, false);
    });
    trChart.reflow();
    trChart.redraw();
  }

  /* ---- Phenology ---- */
  var phChart = (g.charts && g.charts.phenology) || getChart("zg-phenology");
  if (phChart) { g.charts = g.charts||{}; g.charts.phenology = phChart; }
  if (phChart) {
    var phData = ph
      .map(function(d) { return [d.year, d.peak_week]; })
      .sort(function(a, b) { return a[0] - b[0]; });
    phChart.series[0].setData(phData, false);
    var pr  = (g.phenoReg || []).find(function(d) { return d.scientificName === taxon; });
    var prd = pr ? [[pr.x_min, pr.y_min], [pr.x_max, pr.y_max]] : [];
    phChart.series[1].setData(prd, false);
    phChart.reflow();
    phChart.redraw();
  }
}

/* Reflow charts when Bootstrap tabs become visible (Highcharts can not measure
   a hidden container, so the first reveal needs an explicit reflow) */
document.addEventListener("shown.bs.tab", function() {
  setTimeout(function() {
    ["zg-heatmap", "zg-trend", "zg-phenology"].forEach(function(id) {
      try { var c = getChart(id); if (c) { c.reflow(); c.redraw(); } } catch(e2) {}
    });
  }, 100);
});

/* ---- Initialisation ---- */
/* Called once the DOM is ready (readyState guard handles the case where the
   script runs after DOMContentLoaded has already fired, which is common in
   Quarto embed-resources output).
   Data is MERGED into window.__zoogon__ so any charts already registered by
   onRender callbacks are preserved.                                          */
function zoogonInit() {
  var raw = document.getElementById("zoogon-data");
  if (!raw) return;
  var parsed = JSON.parse(raw.textContent);
  if (!window.__zoogon__) window.__zoogon__ = {};
  var g = window.__zoogon__;
  /* merge parsed fields, keep any existing g.charts */
  ["yearCats","yearIdx","heatmap","trend","trendReg",
   "phenology","phenoReg","meta"].forEach(function(k) { g[k] = parsed[k]; });

  var sel = document.getElementById("taxon-select");
  if (!sel) return;
  /* populate only once */
  if (sel.options.length === 0) {
    g.meta.forEach(function(t) {
      var o = document.createElement("option");
      o.value = t.scientificName; o.textContent = t.scientificName;
      sel.appendChild(o);
    });
  }
  /* attach JS event listeners (more reliable than HTML attribute handlers) */
  document.getElementById("taxon-search")
          .addEventListener("input", zoogonFilter);
  sel.addEventListener("change", function() { zoogonUpdate(this.value); });

  /* trigger initial update with first taxon that has heatmap data */
  if (g.meta && g.meta.length) {
    var taxaWithData = new Set((g.heatmap || []).map(function(d){ return d.scientificName; }));
    var firstWithData = g.meta.find(function(t){ return taxaWithData.has(t.scientificName); });
    var initTaxon = firstWithData ? firstWithData.scientificName : g.meta[0].scientificName;
    var sel = document.getElementById("taxon-select");
    if (sel && sel.options.length > 0) sel.value = initTaxon;
    zoogonUpdate(initTaxon);
  }
}

if (document.readyState === "loading") {
  document.addEventListener("DOMContentLoaded", zoogonInit);
} else {
  zoogonInit();
}
'))
```

---

# Dataset Overview

All `r if (exists("taxa_meta")) nrow(taxa_meta) else "–"` taxa in the 40-year
dataset. Click a column header to sort.

```{r dataset-table}
#| echo: false
if (!has_data) {
  htmltools::p("Data not available.", style = "color:#888; font-style:italic")
} else {
  
  # 1. Transform the raw URLs into clickable HTML links
  # Using target="_blank" ensures the link opens in a new tab
  taxa_meta <- taxa_meta |>
    dplyr::mutate(
      link = ifelse(
        is.na(link) | link == "", 
        NA, 
        paste0('<a href="', link, '" target="_blank">View Link</a>') 
        # Note: Replace "View Link" with `link` if you want to display the full URL text instead
      )
    )

  # 2. Render the datatable with escape = FALSE
  DT::datatable(
    taxa_meta,
    colnames  = c(
      "Scientific Name" = "scientificName",
      "Link"            = "link",
      "Class"           = "rankClass",
      "Order"           = "rankOrder"
    ),
    options   = list(pageLength = 25, order = list(list(0, "asc"))),
    rownames  = FALSE,
    escape    = FALSE  # Critical: tells DT to render the HTML rather than printing it as text
  )
}
```

---

# Automated Data Pipeline

This report is produced by a fully automated pipeline running daily at 00:00 UTC
via [GitHub Actions](https://github.com/ioledc/ZOOGoN-40Y/actions). HTML
artifacts are retained for 90 days and are accessible directly through the
Actions UI. All survey data are simultaneously converted into Darwin Core–compliant
format for integration with OBIS, EMODnet Biology, and GBIF.
