---
title: "Gulf of Naples - 40 Years of Zooplankton Biodiversity Assessment (ZOOGoN-40Y)"
author: "Andrea Marcucci (andrea.marcucci@szn.it)"
date: today
format:
  html:
    theme: Darkly
    toc: true
    toc-float: true
    number-sections: true
    code-fold: true
    code-tools: true
    self-contained: true
---

# Aim
This report presents a spatio-temporal analysis of zooplankton abundance patterns at the MareChiara Long-Term Ecological Research (LTER) station in the Gulf of Naples, Mediterranean Sea, covering data recorded between 1984 and 2024. MareChiara is one of the longest-running marine monitoring stations in the Mediterranean, providing a rare 40-year window into plankton community dynamics.

# Data Sources and Manipolation 
The three temporal datasets used in this report covering 1984–2015, 2016–2020, and 2021–present; were obtained from the "Gulf of Naples - 40 Years of Zooplankton Biodiversity Assessment" project, funded by Flanders Marine Institute (VLIZ) and Stazione Zoologica Anton Dohrn (SZN). The legacy dataset (1984–2020), previously maintained on local computers, have been migrated to centralized cloud storage on SharePoint to ensure reproducibility and accessibility. From 2021 onwards, data collection relies on [KoBoToolbox](https://www.kobotoolbox.org) digital survey forms, which replaced the previous local workflow.

Each dataset contains taxonomic identification, life stage classification, and abundance values (ind m⁻³) for zooplankton taxa. The three datasets were merged into a single time series, and copepod records were filtered for analysis. A methodological change occurred in 2021 with the adoption of KoBoToolbox, which introduced the `isCopepod` variable (yes/no) to facilitate taxonomic filtering. To ensure consistency across the entire time series, taxonomic classification variables were harmonized for the 1984–2020 datasets using the `worrms` package. Data manipulation and summarization were performed using `dplyr`, while visualizations were created with `ggplot2`, `cowplot`, and `viridis`.

The code below demonstrates the data loading process and displays the first 10 rows of the final dataset after merge with [WoRMS](https://www.marinespecies.org/) taxonomic data:
```{r}
#| echo: true
#| code-fold: true
# require needed packages
pacman::p_load(
  magrittr,
  scales,
  ggsci,
  lubridate,
  tidyverse,
  viridis,
  cowplot,
  worrms,
  stringr,
  vegan,
  arrow
)

# load package and configurations
library(ZooGoN)
conf <- ZooGoN::read_config()

# load legacy data from SharePoint
legacy_84_15 <- ZooGoN::download_sharepoint_file(
  prefix = "McZoo_84-15.parquet",
  options = conf$storage$sharepoint$credentials,
  bucket = conf$storage$sharepoint$buckets$hot_bucket,
  filename = TRUE
)

legacy_16_20 <- ZooGoN::download_sharepoint_file(
  prefix = "McZoo_16-20.parquet",
  options = conf$storage$sharepoint$credentials,
  bucket = conf$storage$sharepoint$buckets$hot_bucket,
  filename = TRUE
)

# load processed survey data from pipeline (2021-24 + futuri)
legacy_21_24 <- ZooGoN::download_sharepoint_file(
  prefix = conf$ingestion$surveys$preprocessed$file_prefix,
  options = conf$storage$sharepoint$credentials,
  bucket = conf$storage$sharepoint$buckets$automation_bucket,
  format = "parquet"
)

# merge into a single databse
legacy_84_24 <-
  dplyr::bind_rows(legacy_84_15, legacy_16_20, legacy_21_24)

# add Class and Order from Worms
# extrac AphiaID to lsid
aphiaid <-
  legacy_84_24 |>
  dplyr::mutate(
    aphiaID = as.integer(stringr::str_extract(lsid, "[0-9]+"))
  )

# Extract unique AphiaIDs to avoid duplicate WoRMS API calls
ids <- unique(aphiaid$aphiaID)

# get WORms taxonomy
get_class_order <- function(id) {
  tryCatch(
    {
      cls <- wm_classification(id)
      # extract scientificname where rank matches, return NA if not found
      class_val <- cls$scientificname[cls$rank == "Class"]
      order_val <- cls$scientificname[cls$rank == "Order"]
      data.frame(
        aphiaID = id,
        class = if (length(class_val) > 0) class_val[1] else NA_character_,
        order = if (length(order_val) > 0) order_val[1] else NA_character_,
        stringsAsFactors = FALSE
      )
    },
    error = function(e) {
      data.frame(
        aphiaID = id,
        class = NA_character_,
        order = NA_character_,
        stringsAsFactors = FALSE
      )
    }
  )
}

ranks_df <- bind_rows(lapply(ids, get_class_order))

# merge
zoo_84_24 <-
  aphiaid |>
  dplyr::left_join(ranks_df, by = "aphiaID") |>
  dplyr::select(-aphiaID) |>
  dplyr::relocate(class, order, .before = scientificName) |>
  dplyr::rename(rankCalss = class, rankOrder = order)

head(zoo_84_24, 10)
```

# Methods
## Automated Processing Pipeline
Data are processed through the [ZooGoN](https://github.com/ioledc/ZOOGoN-40Y) R package, which implements a fully automated pipeline structured in two parallel way, both triggered daily via [GitHub Actions](https://github.com/ioledc/ZOOGoN-40Y/actions/runs/22932728351). 
The complete pipeline workflow below: 
```{r}
#| echo: false
library(DiagrammeR)
grViz(
  '
  digraph pipeline {
    rankdir=LR
    node [shape=box, style=filled, fillcolor=black, fontcolor=white, fontname=Helvetica]
    edge [color=black, minlen=2]
    
    A [label="Build R Container"]
    B [label="Ingest MC Kobo Surveys"]
    C [label="Process MC Kobo Surveys"]
    D [label="Merge Tidy Analysis-Ready Dataset"]
    E [label="Convert to Darwin Core Format"]
    F [label="Render MC Survey Report"]
    G [label="Build Darwin Core Archive"]
    
    A->B->C->E->G
    C->D->F
  }
'
)
```

## Analysis & Report
- `ingest_surveys()` connects to the KoBoToolbox API, downloads new survey submissions with pagination support, flattens nested JSON data into tabular format, and archives the raw data on SharePoint as CSV and Parquet files. 
- `preprocess_surveys()` cleans and transforms raw survey data, computes abundance values (ind m⁻³) by sex and life stage, and harmonizes taxonomic identifiers against WoRMS via AphiaID.
- `raw_to_tidy()` merges the legacy datasets (1984–2020) with the ongoing survey landings (2021–present) into a single long-format analysis-ready dataset, uploaded to SharePoint as CSV and Parquet.
- `render_report()` downloads the latest processed dataset from SharePoint and renders this report in HTML and PDF formats.

## Darwin Core Archive & GBIF Publication
The function `raw_to_dc()` converts the merged dataset into [Darwin Core](https://dwc.tdwg.org/) format following [OBIS](https://obis.org/) and [EMODnet Biology](https://www.emodnet-biology.eu/) standards, producing three extension tables:

1. Event: sampling metadata including coordinates (40.81°N, 14.25°E), depth range (0–50 m), and sampling protocol.
2. Occurrence: species occurrences linked to WoRMS LSIDs.
3. eMoF: quantitative measurements (ind m⁻³) with standardized [BODC NERC](https://vocab.nerc.ac.uk/) vocabulary terms for sex and life stage.

The function `dc_to_archive()` packages the Darwin Core tables into a Darwin Core Archive (DwC-A) zip file with EML metadata, and uploads it to SharePoint for subsequent publication to [GBIF](https://www.gbif.org/).

The full source code of the ZooGoN pipeline is available on [GitHub](https://github.com/ioledc/ZOOGoN-40Y/tree/main).

## Example of Darwin Core Output 
```{r}
#| echo: false
dc_list <- ZooGoN::download_sharepoint_file(
  prefix = conf$ingestion$surveys$darwincore$file_prefix,
  options = conf$storage$sharepoint$credentials,
  bucket = conf$storage$sharepoint$buckets$automation_bucket,
  format = "rds"
)

knitr::kable(head(dc_list$event, 10), caption = "Event table")
knitr::kable(head(dc_list$occurrence, 10), caption = "Occurrence table")
knitr::kable(head(dc_list$emof, 10), caption = "eMoF table")
```

# Results
The analysis employs two complementary visualization approaches to characterize temporal patterns and detect long-term trends in zooplankton abundance at the LTER-MareChiara station:

1. **Heatmap visualization**: a multi-panel display showing weekly abundance patterns across years, integrated with seasonal profiles and annual anomalies to identify recurrent phenological cycles and interannual variability in community composition.

2. **Trend analysis**: a quarterly-stratified approach based on anomaly-derived regression models, designed to isolate and quantify long-term directional trends while accounting for the confounding effects of natural seasonal variability.

Prior to visualization, abundance data were aggregated by week and year to optimize computational performance and reduce noise associated with individual sampling events. Abundance values were log-transformed (log(x + 1)) to reduce positive skewness and improve the dynamic range of visualizations across taxa spanning several orders of magnitude. These analytical approaches enable the detection of phenological shifts, regime changes, and season-specific trends in zooplankton abundance over the 40-year monitoring period.

The first 10 rows of the zoo_heatmap dataset: 
```{r}
#| echo: true
#| code-fold: true
zoo_heatmap <-
  zoo_84_24 |>
  dplyr::mutate(
    week = lubridate::week(eventDate), # Extract week (1-52)
    year = lubridate::year(eventDate) # Extract year
  ) |>
  dplyr::group_by(year, week, rankCalss, rankOrder, scientificName) |> # Group by year, week, species and life stage
  dplyr::summarise(
    abundance = sum(individualCount, na.rm = TRUE),
    .groups = "drop"
  ) |> # Sum abundances
  dplyr::mutate(log_abundance = log(abundance + 1)) # Log-transformation (+1 to avoid log(0))

head(zoo_heatmap, 10)
```

## Heatmap
The heatmap plots illustrates the abundance across different weeks and years. The x-axis represents the weeks of the year, ranging from 1 to 52, while the y-axis spans several years. Each cell in the heatmap indicates the abundance in individuals per cubic metre, with a color gradient from blue (low abundance) to red (high abundance), as indicated by the legend at the bottom. At the top of the heatmap, a line plot displays the average weekly abundance over the entire period, highlighting the seasonality of abundance. A red point marks the mean maximum abundance, while another red point indicates the first non-zero mean value, providing insight into the average abundance range of the species. On the right side of the heatmap, a side plot displays the mean annual deviation in percentage from the overall annual mean. In `plot_cop_heatmap_complex()` abundance values are log-transformed by default (log = TRUE) in order to reduce skewness and improve the visualization of low and high abundance values.The log-transformation can be disabled by setting log = FALSE. 
```{r}
#| echo: true
#| code-fold: true
# heatmap function
plot_heatmap_complex <- function(
  species_name = NULL,
  rank_class = NULL,
  rank_order = NULL,
  log = TRUE
) {
  # apply taxonomic filters first (if provided), then aggregate or filter by species
  data <-
    zoo_heatmap |>
    dplyr::filter(
      if (!is.null(rank_class)) rankCalss == rank_class else TRUE, # filter by class if provided
      if (!is.null(rank_order)) rankOrder == rank_order else TRUE, # filter by order if provided
      if (!is.null(species_name) && species_name != "All") {
        scientificName == species_name
      } else {
        TRUE
      } # filter by species if provided
    )

  # aggregate all species if species_name is NULL or "All"
  if (is.null(species_name) || species_name == "All") {
    data <-
      data |>
      dplyr::group_by(year, week) |>
      dplyr::summarise(
        abundance = sum(abundance, na.rm = TRUE),
        .groups = "drop"
      ) |>
      dplyr::mutate(
        abundance_plot = if (log) log(abundance + 1) else abundance
      )
  } else {
    # filter by specific species
    data <-
      data |>
      dplyr::mutate(
        abundance_plot = if (log) log_abundance else abundance
      )
  }

  # build dynamic title based on what is being plotted
  plot_title <- if (!is.null(species_name) && species_name != "All") {
    species_name
  } else if (!is.null(rank_order)) {
    rank_order
  } else if (!is.null(rank_class)) {
    rank_class
  } else {
    "All Zooplankton"
  }

  # legend text and theme
  legend_text <- expression(bold(ind ~ m^-3))

  legend_theme <- ggplot2::theme(
    legend.position = "bottom",
    legend.key.width = unit(0.5, "cm"),
    legend.direction = "horizontal",
    legend.key.size = unit(0.2, "cm"),
    legend.title = ggplot2::element_text(size = 10, face = "bold"),
    legend.text = ggplot2::element_text(size = 10, face = "bold")
  )

  # dynamic breaks: visually centered on pseudo-log scale regardless of data range
  min_val <- min(data$abundance_plot[data$abundance_plot > 0], na.rm = TRUE)
  max_val <- max(data$abundance_plot, na.rm = TRUE)

  # compute midpoint on the transformed scale, then back-transform
  sigma <- 5
  mid_transformed <- (asinh(min_val / sigma) + asinh(max_val / sigma)) / 2
  mid_val <- round(sinh(mid_transformed) * sigma) # back-transform to original scale

  breaks_vals <- c(0, mid_val, round(max_val))

  # main heatmap: weeks (x) vs years (y), colored by abundance
  heatmap_plot <-
    data |>
    ggplot2::ggplot(ggplot2::aes(x = week, y = year, fill = abundance_plot)) +
    ggplot2::geom_tile(color = "grey90") +
    ggplot2::theme_bw() +
    ggplot2::scale_fill_gradientn(
      colours = c("blue", "cyan", "green", "yellow", "orange", "red"),
      trans = scales::pseudo_log_trans(sigma = 5),
      breaks = breaks_vals, # dynamic breaks adapted to the plotted data
      labels = scales::label_number(),
      na.value = "grey90"
    ) +
    ggplot2::coord_cartesian(expand = FALSE) +
    ggplot2::scale_x_continuous(breaks = seq(1, 52, by = 5)) +
    ggplot2::scale_y_continuous(breaks = seq(1984, 2024, by = 4)) +
    ggplot2::theme(
      panel.background = ggplot2::element_blank(),
      panel.grid = ggplot2::element_blank(),
      legend.box.just = "left",
      legend.box.margin = margin(t = 0, r = 300, b = 0, l = 0),
      axis.text.x = ggplot2::element_text(size = 12, face = "bold"),
      axis.text.y = ggplot2::element_text(size = 12, face = "bold"),
      axis.title.x = ggplot2::element_text(size = 16, face = "bold"),
      axis.title.y = ggplot2::element_text(size = 16, face = "bold"),
      plot.title = ggplot2::element_text(size = 16, face = "bold")
    ) +
    legend_theme +
    ggplot2::labs(
      x = "Weeks",
      y = "Years",
      fill = legend_text
    ) +
    guides(alpha = "none")

  # top marginal plot: mean abundance by week across all years
  line <-
    data |>
    dplyr::group_by(week) |>
    dplyr::summarise(abundance_plot = mean(abundance_plot, na.rm = TRUE))

  # identify min and max values for annotation
  minmax <-
    line |>
    dplyr::mutate(
      label_row = dplyr::case_when(
        abundance_plot == min(abundance_plot, na.rm = TRUE) ~ "min",
        abundance_plot == max(abundance_plot, na.rm = TRUE) ~ "max",
        TRUE ~ NA_character_
      )
    ) |>
    dplyr::filter(!is.na(label_row)) |>
    dplyr::arrange(abundance_plot)

  line_plot <- ggplot2::ggplot() +
    ggplot2::theme_void() +
    ggplot2::stat_smooth(
      data = line,
      ggplot2::aes(week, abundance_plot),
      method = "loess",
      color = "#d64d4d",
      se = FALSE,
      span = 0.25,
      linewidth = 0.5
    ) +
    ggplot2::geom_area(
      data = line,
      ggplot2::aes(week, abundance_plot),
      fill = "grey95",
      color = "grey70"
    ) +
    ggplot2::geom_point(
      data = line,
      ggplot2::aes(week, abundance_plot, color = abundance_plot)
    ) +
    ggplot2::geom_text(
      data = minmax,
      ggplot2::aes(
        x = week,
        y = abundance_plot,
        label = round(abundance_plot, 1),
        vjust = ifelse(label_row == "max", 1.5, -0.5)
      ),
      color = "firebrick",
      size = 6
    ) +
    ggplot2::geom_point(
      data = minmax,
      ggplot2::aes(x = week, y = abundance_plot),
      color = "firebrick"
    ) +
    ggplot2::scale_color_gradientn(
      colours = c("blue", "cyan", "green", "yellow", "orange", "red")
    ) +
    ggplot2::scale_x_continuous(limits = c(1, 52), expand = c(0, 0)) +
    ggplot2::coord_cartesian(xlim = c(1, 52), expand = FALSE, clip = "off") +
    ggplot2::theme(
      legend.position = "none",
      plot.margin = margin(t = 15, b = 5, l = 70, r = 69)
    )

  # right side panel: annual anomalies as % deviation from mean
  side_tile <-
    data |>
    dplyr::group_by(year) |>
    dplyr::summarise(abundance_plot = mean(abundance_plot, na.rm = TRUE)) |>
    dplyr::mutate(
      mean_abundance = mean(abundance_plot, na.rm = TRUE),
      mean_anomaly = (abundance_plot - mean_abundance) / mean_abundance * 100,
      mean_anomaly_scale = pmax(pmin(mean_anomaly, 100), -100)
    )

  side_plot <-
    side_tile |>
    ggplot2::ggplot() +
    ggplot2::geom_tile(
      ggplot2::aes(x = "group", y = year, fill = mean_anomaly_scale),
      alpha = 0.6,
      color = "grey90"
    ) +
    ggplot2::geom_text(
      ggplot2::aes(
        x = "group",
        y = year,
        label = ifelse(
          !is.na(mean_anomaly),
          paste0(round(mean_anomaly, 0), "%"),
          "-"
        )
      ),
      size = 2,
      color = "grey30"
    ) +
    ggplot2::theme_void() +
    ggplot2::scale_fill_gradient2(
      low = "#56a8a9",
      mid = "white",
      high = "#a95756",
      midpoint = 0,
      na.value = "white"
    ) +
    ggplot2::scale_y_continuous(breaks = seq(1984, 2024, by = 4)) +
    ggplot2::coord_cartesian(expand = FALSE) +
    ggplot2::labs(fill = "Mean difference (%)") +
    ggplot2::theme(
      plot.margin = margin(t = 0, r = 12, b = 0, l = -3.5),
      panel.background = ggplot2::element_blank(),
      panel.grid = ggplot2::element_blank(),
      legend.box.just = "right",
      legend.box.margin = margin(t = 0, r = 0, b = 0, l = -250)
    ) +
    legend_theme +
    guides(alpha = "none")

  # combine panels
  p1 <- cowplot::plot_grid(
    heatmap_plot,
    side_plot,
    ncol = 2,
    rel_widths = c(12, 1),
    align = "h"
  )

  combined <- cowplot::plot_grid(line_plot, p1, nrow = 2, rel_heights = c(1, 5))

  # add title
  title <- cowplot::ggdraw() +
    cowplot::draw_label(
      bquote(italic(.(plot_title))), # dynamic title based on what is being plotted
      fontface = 'bold',
      size = 20,
      x = 0,
      hjust = 0
    ) +
    ggplot2::theme(
      plot.margin = margin(0, 0, 0, 7)
    )

  # return final plot
  cowplot::plot_grid(
    title,
    combined,
    ncol = 1,
    rel_heights = c(0.05, 1)
  )
}
```

## Temporal trend
The temporal trend plot illustrates the mean deviation in abundance over time, featuring a linear regression line and the p-value of the regression for each quarter of the year. The x-axis represents the date, and the y-axis shows the monthly mean deviation in abundance. This deviation is calculated by subtracting the overall mean for each specific month from the corresponding monthly values in the series. For example, each January value is subtracted from the overall January mean, each February value from the overall February mean, and so on. The `plot_temporal_trend()` function performs deseasonalized anomaly analysis with quarterly stratification to detect long-term trends.
```{r}
#| echo: true
#| code-fold: true
# trend function
plot_temporal_trend <- function(
  species_name = NULL,
  rank_class = NULL,
  rank_order = NULL
) {
  # apply taxonomic filters first (if provided)
  zoo_filtered <-
    zoo_84_24 |>
    dplyr::filter(
      if (!is.null(rank_class)) rankCalss == rank_class else TRUE, # filter by class if provided
      if (!is.null(rank_order)) rankOrder == rank_order else TRUE, # filter by order if provided
      if (!is.null(species_name) && species_name != "All") {
        scientificName == species_name
      } else {
        TRUE
      } # filter by species if provided
    )

  # aggregate all species if species_name is NULL or "All"
  if (is.null(species_name) || species_name == "All") {
    filtered_data <-
      zoo_filtered |>
      dplyr::group_by(eventDate) |>
      dplyr::summarise(
        abundance = sum(individualCount, na.rm = TRUE),
        .groups = "drop"
      )
  } else {
    filtered_data <-
      zoo_filtered |>
      dplyr::filter(scientificName == species_name) |>
      dplyr::group_by(eventDate) |>
      dplyr::summarise(
        abundance = sum(individualCount, na.rm = TRUE),
        .groups = "drop"
      )
  }

  # build dynamic title based on what is being plotted
  plot_title <- if (!is.null(species_name) && species_name != "All") {
    species_name
  } else if (!is.null(rank_order)) {
    rank_order
  } else if (!is.null(rank_class)) {
    rank_class
  } else {
    "All Zooplankton"
  }

  # calculate deseasonalized anomalies with quarterly stratification
  data_df <-
    filtered_data |>
    dplyr::mutate(
      month = lubridate::month(eventDate),
      quarter = lubridate::quarter(eventDate), # Q1-Q4 for seasonal analysis
      year = lubridate::year(eventDate),
      month_date = lubridate::floor_date(eventDate, "month"),
      series = ifelse(year < 1993, "before", "after") # temporal split
    ) |>
    dplyr::group_by(series, month) |>
    dplyr::mutate(
      mean_abundance = mean(abundance),
      anomaly = abundance - mean_abundance # remove seasonal cycle
    ) |>
    dplyr::group_by(series, month_date) |>
    dplyr::summarise(
      quarter = dplyr::first(quarter),
      month = dplyr::first(month),
      anomaly = mean(anomaly),
      .groups = "drop"
    ) |>
    dplyr::mutate(
      quarter = as.character(quarter)
    ) |>
    dplyr::ungroup()

  # split into before/after periods
  data_df <- split(data_df, data_df$series)

  # linear regression by quarter (post-1993 only)
  trend_res <-
    data_df$after |>
    dplyr::group_by(quarter) |>
    dplyr::do(broom::tidy(lm(anomaly ~ month_date, data = .))) |>
    dplyr::filter(term == "month_date") |>
    dplyr::select(quarter, estimate, p.value)

  # format statistical results as text annotations
  annotations_text <-
    trend_res |>
    dplyr::arrange(quarter) |>
    dplyr::mutate(
      label = sprintf("Q%s: %.4f, p=%.4f", quarter, estimate, p.value)
    ) |>
    dplyr::pull(label) |>
    paste(collapse = "\n")

  # time series plot with quarterly color coding
  ggplot2::ggplot() +
    ggplot2::theme_minimal() +
    # points for "before" period (pre-1993)
    ggplot2::geom_point(
      data = data_df$before,
      ggplot2::aes(month_date, anomaly, color = quarter, alpha = abs(anomaly))
    ) +
    # points for "after" period (post-1993)
    ggplot2::geom_point(
      data = data_df$after,
      ggplot2::aes(month_date, anomaly, color = quarter, alpha = abs(anomaly))
    ) +
    # connecting lines
    ggplot2::geom_line(
      data = data_df$before,
      ggplot2::aes(month_date, anomaly),
      alpha = 0.3,
      linewidth = 0.1
    ) +
    ggplot2::geom_line(
      data = data_df$after,
      ggplot2::aes(month_date, anomaly),
      alpha = 0.3,
      linewidth = 0.1
    ) +
    # zero reference line (climatological mean)
    ggplot2::geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
    # linear trend lines by quarter (post-1993 only)
    ggplot2::stat_smooth(
      data = data_df$after,
      ggplot2::aes(month_date, anomaly, group = quarter, color = quarter),
      method = "lm",
      se = FALSE,
      linewidth = 0.8
    ) +
    # JAMA color palette for quarterly distinction
    ggsci::scale_color_jama() +
    ggplot2::scale_x_date(
      date_breaks = "4 years",
      labels = scales::label_date(format = "%b\n%Y", locale = "en_US.UTF-8")
    ) +
    ggplot2::coord_cartesian(expand = FALSE) +
    # statistical results annotation
    ggplot2::annotate(
      "text",
      x = as.Date("1985-12-01"),
      y = max(data_df$after$anomaly, na.rm = TRUE),
      label = annotations_text,
      hjust = 0,
      vjust = 1,
      color = "grey40"
    ) +
    ggplot2::theme(legend.position = "bottom") +
    ggplot2::guides(alpha = "none") +
    ggplot2::labs(
      title = bquote(italic(.(plot_title))), # dynamic title based on what is being plotted
      x = "Date",
      y = "Monthly mean deviation\n(Abundance)",
      color = "Annual quarter"
    )
}
```

## Temporal trend
The temporal trend plot illustrates the mean deviation in abundance over time, featuring a linear regression line and the p-value of the regression for each quarter of the year. The x-axis represents the date, and the y-axis shows the monthly mean deviation in abundance. This deviation is calculated by subtracting the overall mean for each specific month from the corresponding monthly values in the series. For example, each January value is subtracted from the overall January mean, each February value from the overall February mean, and so on.

# Data Visualization Examples 
The following code illustrates how to call the two main visualization functions at different taxonomic levels: 
```{r}
#| echo: true
#| code-fold: true
plot_heatmap_complex(species_name = "All", log = F) # all zooplankton
plot_heatmap_complex(rank_class = "Copepoda", log = F) # single class
plot_heatmap_complex(rank_order = "Onychopoda", log = F) # single order
plot_heatmap_complex(species_name = "Acartia clausi", log = F) # single sepcies
```

```{r}
#| echo: true
#| code-fold: true
plot_temporal_trend(species_name = "All") # all zooplankton
plot_temporal_trend(rank_class = "Thecostraca") # single class
plot_temporal_trend(rank_order = "Euphausiacea") # single order
plot_temporal_trend(species_name = "Acartia clausi") # single species
```

 





# Darwin Core Archive & GBIF Publication
The function `raw_to_dc()` converts the merged dataset into [**Darwin Core**](https://dwc.tdwg.org/) format following [OBIS](https://obis.org/) and [EMODnet Biology](https://www.emodnet-biology.eu/) standards, producing three extension tables:

1. Event: sampling metadata including coordinates (40.81°N, 14.25°E), depth range (0–50 m), and sampling protocol.
2. Occurrence: species occurrences linked to WoRMS LSIDs.
3. eMoF: quantitative measurements (ind m⁻³) with standardized [BODC NERC](https://vocab.nerc.ac.uk/) vocabulary terms for sex and life stage.

The function `dc_to_archive()` packages the Darwin Core tables into a Darwin Core Archive (DwC-A) zip file with EML metadata, and uploads it to SharePoint for subsequent publication to [**GBIF**](https://www.gbif.org/).

# Example of Darwin Core Output 
```{r}
#| echo: false
dc_list <- ZooGoN::download_sharepoint_file(
  prefix = conf$ingestion$surveys$darwincore$file_prefix,
  options = conf$storage$sharepoint$credentials,
  bucket = conf$storage$sharepoint$buckets$automation_bucket,
  format = "rds"
)

knitr::kable(head(dc_list$event, 10), caption = "Event table")
knitr::kable(head(dc_list$occurrence, 10), caption = "Occurrence table")
knitr::kable(head(dc_list$emof, 10), caption = "eMoF table")
```

# Report Automation
This report is fully automated and self-regenerating. The pipeline runs every day at 00:00 UTC, ensuring that new field data collected via KoBoToolbox are reflected in the report the following morning. The rendered HTML and PDF outputs are archived as downloadable artifacts on GitHub Actions for 90 days.