---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ZooGoN <img src="man/figures/logo.svg" align="right" height="139"/>

<!-- badges: start -->

[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental) [![R-CMD-check](https://github.com/ioledc/ZOOGoN-40Y/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/ioledc/ZOOGoN-40Y/actions/workflows/R-CMD-check.yaml)

<!-- badges: end -->

**Gulf of Naples - 40 Years of Zooplankton Biodiversity Assessment**

ZooGoN standardizes taxonomic names in Mediterranean zooplankton datasets spanning four decades (1984-2024) from the LTER-MareChiara station in the Gulf of Naples. This R package is part of the [**DTO-BioFlow project**](https://dto-bioflow.eu) (Digital Twin Ocean - Biodiversity Flow Integration) under the EU Horizon Mission "Restore our Ocean & Waters by 2030".

## Project Context

This package processes the most comprehensive long-term zooplankton dataset from the Western Mediterranean Sea, including:

-   📊 **1,506 zooplankton samples** (1984-2024)\
-   🦐 **148 copepod species** + 61 other taxa
-   📍 **LTER-MareChiara station**, Gulf of Naples, Tyrrhenian Sea
-   🏛️ **Stazione Zoologica Anton Dohrn** (SZN)
-   🌍 Integration with **European Digital Twin of the Ocean**
-   💰 €60,000 DTO-BioFlow FSTP grant funding

## Key Features

-   **Taxonomic Standardization**: Robust genus-species extraction from complex taxonomic strings
-   **Species Complex Handling**: Converts multi-species entries (e.g., "Sardinella+Sardinops" → "Sardinella spp")
-   **Family-level Processing**: Creates standardized names from family entries (e.g., "Clupeidae n.i." → "Clupegenus sp")
-   **Darwin Core Compliance**: Converts datasets to international biodiversity standards
-   **FAIR Data Principles**: Ensures Findable, Accessible, Interoperable, Reusable data
-   **EMODnet Biology Integration**: Quality-controlled data publication workflow

## Installation

You can install the development version of ZooGoN from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("ioledc/ZOOGoN-40Y")
```

## Usage

### Basic Taxonomic Standardization

```{r example, eval = FALSE}
library(ZooGoN)

# Example taxonomic names from Gulf of Naples zooplankton samples
taxa_examples <- c(
  "Sardinella+Sardinops",                    # Species complex
  "Clupeidae n.i.",                          # Family level  
  "Engraulis - group",                       # Higher group
  "Lutjanus (Paradies) argentimaculatus (Forsskål, 1775)",  # Full binomial
  "Chiridius poppei Giesbrecht, 1893"       # Standard binomial
)

# Standardize taxonomic names
standardized <- extract_genus_species(taxa_examples)
print(standardized)

# Expected output:
#   original_name                                          genus_species
#   <chr>                                                  <chr>        
# 1 Sardinella+Sardinops                                   Sardinella spp
# 2 Clupeidae n.i.                                         Clupegenus sp
# 3 Engraulis - group                                      Engraulis indet
# 4 Lutjanus (Paradies) argentimaculatus (Forsskål, 1775) Lutjanus argentimaculatus
# 5 Chiridius poppei Giesbrecht, 1893                     Chiridius poppei
```

### Data Processing Workflow

The package implements a complete data processing pipeline for the LTER-MareChiara dataset:

```{r workflow, eval = FALSE}
# Load and process zooplankton abundance data
# This workflow processes 40 years of Gulf of Naples data

# 1. Read raw Excel files
raw_data <- readxl::read_xlsx("data/lter_zoo_84_13.xlsx")
sample_ids <- readxl::read_xlsx("data/ids.xlsx")

# 2. Apply taxonomic standardization
standardized_data <- raw_data %>%
  mutate(
    standardized_taxa = extract_genus_species(TAXA)$genus_species
  )

# 3. Convert to Darwin Core format
# Creates event, occurrence, and eMoF (extended Measurement or Fact) tables
# following international biodiversity data standards
```

## Dataset Overview

The LTER-MareChiara zooplankton dataset represents one of the longest continuous time series in the Mediterranean Sea:

| Period | Frequency | Samples | Net Type | Fixation |
|---------------|---------------|---------------|---------------|---------------|
| 1984-1990 | Biweekly | 156 | Indian Ocean (200μm, 113cm) | Formaldehyde 2-4% |
| 1991-1994 | *Interruption* | \- | \- | \- |
| 1995-2015 | Weekly | 1,092 | Indian Ocean (200μm, 113cm) | Formaldehyde 2-4% |
| 2016-2024 | Weekly | 258 | WP2 (200μm, 70cm) | Ethanol 96% |

**Total: 1,506 samples • 148 copepod species • 61 other taxa**

## Data Standards & Compliance

ZooGoN ensures compatibility with international biodiversity data standards:

-   **🗂️ Darwin Core Archive**: International standard for biodiversity data
-   **🌐 WoRMS Integration**: World Register of Marine Species taxonomic validation\
-   **📊 BODC NERC Vocabulary**: Standardized measurement terminology
-   **⚡ EMODnet Biology**: European marine biodiversity data infrastructure
-   **🏷️ ISO19115 Metadata**: International metadata standards
-   **📄 FAIR Principles**: Findable, Accessible, Interoperable, Reusable data

## Contributing

This package is part of the DTO-BioFlow project timeline (2025-2026):

-   **May 2025**: Project initiation, Paris workshop
-   **August 2025**: First interim report, EMODnet training completion
-   **December 2025**: Second interim report
-   **April 2026**: Final deliverables and EMODnet Biology publication

## Citation

```{r citation, eval = FALSE}
citation("ZooGoN")
```

## Funding

This work is supported by the DTO-BioFlow project (HORIZON-MISS-2022-OCEAN-01-07) under the EU Mission "Restore our Ocean & Waters by 2030" through a Financial Support to Third Parties (FSTP) grant of €60,000.

## Contact

- **Principal Investigator**: Iole Di Capua (<iole.dicapua@szn.it>)
**Institution**: Stazione Zoologica Anton Dohrn, Naples, Italy
**ORCID**:
  [0000-0003-2959-8977](https://orcid.org/0000-0003-2959-8977)
- **Principal Analyst**: Lorenzo Longobardi (<lorenzo.longobardi@gmail.com>)
**Institution**: WorldFish
**ORCID**:
  [0000-0003-2959-8977](https://orcid.org/0000-0003-3126-7341)
## Acknowledgments

-   LTER-MareChiara research station
-   DTO-BioFlow project consortium\
-   EMODnet Biology data infrastructure
-   European Digital Twin of the Ocean initiative