---
title: "Darwin Core Output for LTER-MareChiara Legacy Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Darwin Core Output for LTER-MareChiara Legacy Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(ZooGoN)
library(dplyr)
```

# Overview

This vignette shows how the package builds Darwin Core tables from the
LTER-MareChiara zooplankton dataset. The conversion is handled by `raw_to_dc()`,
which downloads the merged tidy dataset from SharePoint (produced by
`raw_to_tidy()`), adds the required Darwin Core fields, and uploads the
resulting tables back to SharePoint as a versioned RDS file. The function uses
data that already contains WoRMS LSIDs and does not perform additional taxonomic
queries.

## Prerequisites

- `inst/config.yml` configured with SharePoint credentials and bucket names.
- The tidy dataset must exist in the `automation_bucket` (produced by running
  `raw_to_tidy()` earlier in the pipeline). This merged dataset covers
  1984-2024 and contains preprocessed records with WoRMS LSIDs.
- Optional: set `verbose = FALSE` to silence log messages.

# Expected Input Structure

`raw_to_dc()` expects the tidy dataset to contain at least:

- `eventID`: sampling event identifier (for example `"mc_1"`).
- `eventDate`: sampling date (`Date` column).
- `scientificname`: WoRMS validated taxon name.
- `lsid`: WoRMS Life Science Identifier.
- `individualCount`: abundance measurement (individuals per cubic meter).
- `lifeStage`: life stage codes (`f`, `m`, `j`, `fm`, `fmj`).

Any extra columns are carried through and pivoted into the measurement table.

# Running the conversion

```{r run-raw-to-dc}
# Convert tidy data to Darwin Core format
# Tables are uploaded to SharePoint automatically
raw_to_dc()
```

The function downloads the tidy dataset from SharePoint, builds the three Darwin
Core tables (Event, Occurrence, eMoF), and uploads them as a versioned RDS file
to the `automation_bucket`.

To build a Darwin Core Archive with EML metadata on top of the tables:

```{r build-archive}
# Build DwC-A zip and upload to SharePoint
dc_to_archive()
```

# Darwin Core tables

## Event

`raw_to_dc()` builds one row per unique event with fixed station metadata.

```{r event-glimpse}
# After downloading the DC tables from SharePoint:
# event_table %>%
#   select(eventID, eventDate, decimalLatitude, decimalLongitude, samplingProtocol) %>%
#   head()
```

Columns include:

- `eventID`, `eventDate`
- `decimalLatitude` = 40.81, `decimalLongitude` = 14.25
- `locality`, `country`, `stateProvince`, `waterBody`
- `maximumDepthInMeters` = 50, `minimumDepthInMeters` = 0
- `samplingProtocol`, `sampleSizeValue` (= 1), `sampleSizeUnit` (= "sample")

## Occurrence

Occurrences are derived from the tidy data with an automatically generated
`occurrenceID` and presence flag.

- `occurrenceStatus` is set to `"present"` when `individualCount > 0`, otherwise
  `"absent"`.
- `scientificName` and `scientificNameID` come directly from the input (no new
  validation is run).

## eMoF (Extended Measurement or Fact)

Measurements are produced by pivoting the remaining columns per occurrence.

Mapping logic implemented in `raw_to_dc()`:

- Measurements with values `f`, `m`, `fm`, `fmj` are labelled as
  `measurementType = "sex"` with
  `measurementTypeID = http://vocab.nerc.ac.uk/collection/P01/current/ENTSEX01/`.
- Measurements with value `j` are labelled `measurementType = "lifeStage"` with
  `measurementTypeID = http://vocab.nerc.ac.uk/collection/P01/current/LSTAGE01/`.
- `individualCount` keeps `measurementType = "individualCount"` with
  `measurementTypeID = https://vocab.nerc.ac.uk/collection/S06/current/S0600002/`
  and `measurementUnitID = http://vocab.nerc.ac.uk/collection/P06/current/UPMM/`.
- `measurementValueID` is set for the coded values above (S10/S11 URIs); other
  measurements keep `NA`.

# Processing metadata

`raw_to_dc()` stores processing metadata alongside the tables:

- `processing_info`: event, occurrence and measurement counts, date range, unique taxa count.
- `metadata`: dataset title, contact, institution, license, project tag, and the processing statistics.

# Publishing (optional)

If you need to publish the archive to GBIF:

- Production: supply your real organization/installation keys and a public DwC-A URL to `register_gbif_dataset()`.
- Test: use the GBIF-Test demo helper `register_gbif_dataset_test()` with the demo credentials and a public DwC-A URL to check the flow safely.

# Limitations and next steps

- The function assumes WoRMS identifiers are already present; it does not call
  external taxonomic services.
- No automated quality control is applied beyond field renaming and simple
  presence/absence derivation. Add your own checks before publication.
